---
title: "GRUK_veg_analyse_2021_2024"
author: "Joachim Töpper"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(readxl)
library(sf)
library(plyr)
library(tidyverse)
library(RColorBrewer)
library("gridExtra") 
library(ggridges)
library(ggpubr)
library(hrbrthemes)
library(tmap)
library(tmaptools)
library(raster)
library(stars)
library(betareg)
library(glmmTMB)
library(knitr)

knitr::opts_chunk$set(echo = TRUE)
```

# upload data sets

GRUK data: GRUK.species contains the species data, GRUK.variable contains site data, GRUK2021.condition contains a field-based condition assessment from the 2021 season.
```{r, echo = F, warning = F}
# read data from P:/15211700_gruk_2021_2024/Data
#excel_sheets("P:/15211700_gruk_2021_2024/Data/GRUK_alle_artsdata_2020-24.xlsx")

GRUK.species <- read_excel("P:/15211700_gruk_2021_2024/Data/GRUK_alle_artsdata_2020-24.xlsx",sheet="Arter i ruter")
GRUK.ruter <- read_excel("P:/15211700_gruk_2021_2024/Data/GRUK_alle_artsdata_2020-24.xlsx",sheet="Ruter")
GRUK.sirkler <- read_excel("P:/15211700_gruk_2021_2024/Data/GRUK_alle_artsdata_2020-24.xlsx",sheet="Sirkler")
GRUK.polygoner <- read_excel("P:/15211700_gruk_2021_2024/Data/GRUK_alle_artsdata_2020-24.xlsx",sheet="Polygoner")

```

Plant indicators from Tyler et al. (2021) and Grime (1974) are saved as ind.Tyler and ind.Grime.
```{r, echo = F}
# data from cache
ind.Tyler<-readRDS(paste0(here::here(),"/data/cache/ind.Tyler.RDS"))
ind.Grime<-readRDS(paste0(here::here(),"/data/cache/ind.Grime.RDS"))
```

Generalized species lists (reference communities): natopen_NiN_ref contains the reference species lists, natopen_NiN_ref_spInfo contains additional taxonomic information for each species.
```{r, echo = F}
# data from cache
natopen_NiN_ref<-readRDS(paste0(here::here(),"/data/cache/natopen_NiN_ref.RDS"))
natopen_NiN_ref_spInfo<-readRDS(paste0(here::here(),"/data/cache/natopen_NiN_ref_spInfo.RDS"))
```

# Data handling
- Checking for errors
- Checking species nomenclature in the different species lists to make species and indicator data possible to merge
- Merging indicator data with monitoring data and indicator data with reference data
(not shown here, but documented in the code)



```{r, include = F}

### Plant indicator data
# trimming away sub-species & co, and descriptor info
ind.Grime[,'species.orig'] <- ind.Grime[,'species']
ind.Grime[,'species'] <- word(ind.Grime[,'species'], 1,2)

# dealing with 'duplicates'
ind.Grime[duplicated(ind.Grime[,'species']),"species"]
ind.Grime.dup <- ind.Grime[duplicated(ind.Grime[,'species']),c("species")]
ind.Grime[ind.Grime$species %in% ind.Grime.dup,]
# getting rid of the duplicates
ind.Grime <- ind.Grime %>% filter( !(species.orig %in% list("Carex viridula brachyrrhyncha",
                                                            "Dactylorhiza fuchsii praetermissa",
                                                            "Medicago sativa varia",
                                                            "Montia fontana chondrosperma",
                                                            "Papaver dubium lecoqii",
                                                            "Sanguisorba minor muricata")
) )
#ind.Grime[duplicated(ind.Grime[,'species']),"species"]


names(ind.Tyler)[1] <- 'species'
ind.Tyler[,'species.orig'] <- ind.Tyler[,'species']
ind.Tyler <- ind.Tyler %>% mutate(species = str_remove(species, "sect. "))
ind.Tyler$species <- as.factor(ind.Tyler$species)
#summary(ind.Tyler$species)
ind.Tyler <- ind.Tyler[!is.na(ind.Tyler$species),]

ind.Tyler[,'species'] <- word(ind.Tyler[,'species'], 1,2)


#ind.Tyler2 <- ind.Tyler
#ind.Tyler <- ind.Tyler2
#ind.Tyler[duplicated(ind.Tyler[,'species']),"species"]
ind.Tyler.dup <- ind.Tyler[duplicated(ind.Tyler[,'species']),"species"]
#ind.Tyler[ind.Tyler$species %in% ind.Tyler.dup,c("Light","Moisture","Soil_reaction_pH","Nitrogen","species.orig","species")]
ind.Tyler <- ind.Tyler %>% filter( !(species.orig %in% list("Ammophila arenaria x Calamagrostis epigejos",
                                                            "Anemone nemorosa x ranunculoides",
                                                            "Armeria maritima ssp. elongata",
                                                            "Asplenium trichomanes ssp. quadrivalens",
                                                            "Calystegia sepium ssp. spectabilis",
                                                            "Campanula glomerata 'Superba'",
                                                            "Dactylorhiza maculata ssp. fuchsii",
                                                            "Erigeron acris ssp. droebachensis",
                                                            "Erigeron acris ssp. politus",
                                                            "Erysimum cheiranthoides L. ssp. alatum",
                                                            "Euphrasia nemorosa x stricta var. brevipila",
                                                            "Galium mollugo x verum",
                                                            "Geum rivale x urbanum",
                                                            "Hylotelephium telephium (ssp. maximum)",
                                                            "Juncus alpinoarticulatus ssp. rariflorus",
                                                            "Lamiastrum galeobdolon ssp. argentatum",
                                                            "Lathyrus latifolius ssp. heterophyllus",
                                                            "Medicago sativa ssp. falcata",
                                                            "Medicago sativa ssp. x varia",
                                                            "Monotropa hypopitys ssp. hypophegea",
                                                            "Ononis spinosa ssp. hircina",
                                                            "Ononis spinosa ssp. procurrens",
                                                            "Pilosella aurantiaca ssp. decolorans",
                                                            "Pilosella aurantiaca ssp. dimorpha",
                                                            "Pilosella cymosa ssp. gotlandica",
                                                            "Pilosella cymosa ssp. praealta",
                                                            "Pilosella officinarum ssp. peleteranum",
                                                            "Poa x jemtlandica (Almq.) K. Richt.",
                                                            "Poa x herjedalica Harry Sm.",
                                                            "Ranunculus peltatus ssp. baudotii",
                                                            "Sagittaria natans x sagittifolia",
                                                            "Salix repens ssp. rosmarinifolia",
                                                            "Stellaria nemorum L. ssp. montana",
                                                            "Trichophorum cespitosum ssp. germanicum")
) )
#ind.Tyler[duplicated(ind.Tyler[,'species']),"species"]
ind.Tyler.dup <- ind.Tyler[duplicated(ind.Tyler[,'species']),"species"]
#ind.Tyler[ind.Tyler$species %in% ind.Tyler.dup,c("Light","Moisture","Soil_reaction_pH","Nitrogen","species.orig","species")]

# only hybrids left -> get rid of these
ind.Tyler <- ind.Tyler[!duplicated(ind.Tyler[,'species']),]
#ind.Tyler[duplicated(ind.Tyler[,'species']),"species"]

ind.Tyler$species <- as.factor(ind.Tyler$species)
#summary(ind.Tyler$species)
# no duplicates left

# merge indicator data
ind.dat <- merge(ind.Grime,ind.Tyler, by="species", all=T)
#summary(ind.dat)
#ind.dat[duplicated(ind.dat[,'species']),"species"]
ind.dat$species <- as.factor(ind.dat$species)
#summary(ind.dat$species)
head(ind.dat)



### GRUK data handling
names(GRUK.ruter)
names(GRUK.species)
names(GRUK.sirkler)
names(GRUK.polygoner)
#names(GRUK2021.condition)

## GRUK species data handling
names(GRUK.species)<-c("ObjectID","GlobalID","ParentGlobalID","PolygonID","RuteID","Norsknavn","Latinsknavn","art_dekning","RAkat","CreationDate","Creator","EditDate","Editor","Navn","karplantenavn","karplantedekning","veg_html_row","x")

# fix species names
unique(as.factor(GRUK.species$Latinsknavn))
GRUK.species <- GRUK.species %>%
  mutate(Species=word(Latinsknavn, 1, 2)) # lose subspecies
unique(as.factor(GRUK.species$Species))

# merge species data with indicators
GRUK.species.ind <- merge(x=GRUK.species[,c("Species", "art_dekning", "ParentGlobalID","PolygonID","RuteID")], 
                          y= ind.dat[,c("species","CC", "SS", "RR","Light", "Nitrogen", "Soil_disturbance")],
                          by.x="Species", by.y="species", all.x=T)
summary(GRUK.species.ind)

# checking which species didn't find a match
unique(GRUK.species.ind[is.na(GRUK.species.ind$Light & 
                                is.na(GRUK.species.ind$RR)),'Species'])

# fix species name issues
ind.dat <- ind.dat %>% 
  #  mutate(species=str_replace(species,"Aconitum lycoctonum", "Aconitum septentrionale")) %>% 
  #  mutate(species=str_replace(species,"Carex simpliciuscula", "Kobresia simpliciuscula")) %>%
  #  mutate(species=str_replace(species,"Carex myosuroides", "Kobresia myosuroides")) %>%
  #  mutate(species=str_replace(species,"Artemisia rupestris", "Artemisia norvegica")) %>%
  mutate(species=str_replace(species,"Cotoneaster simonsii", "Cotoneaster symondsii")) %>%
  mutate(species=str_replace(species,"Rosa vosagica", "Rosa vosagiaca"))

GRUK.species <- GRUK.species %>%
  mutate(species=str_replace(Species,"Acinos arvensis", "Clinopodium acinos")) %>%
  mutate(Species=str_replace(Species,"Arabis wahlenbergii", "Arabis hirsuta")) %>%
  #  mutate(Species=str_replace(Species,"Arctous alpinus", "Arctous alpina")) %>%
  #  mutate(Species=str_replace(Species,"Betula tortuosa", "Betula pubescens")) %>%
  #  mutate(Species=str_replace(Species,"Blysmopsis rufa", "Blysmus rufus")) %>%
  #  mutate(Species=str_replace(Species,"Cardamine nymanii", "Cardamine pratensis")) %>%
  #  mutate(Species=str_replace(Species,"Carex adelostoma", "Carex buxbaumii")) %>%
  #  mutate(Species=str_replace(Species,"Carex leersii", "Carex echinata")) %>%
  mutate(Species=str_replace(Species,"Carex paupercula", "Carex magellanica")) %>%
  #  mutate(Species=str_replace(Species,"Carex simpliciuscula", "Kobresia simpliciuscula")) %>%
  mutate(Species=str_replace(Species,"Carex viridula", "Carex flava")) %>%
  #  mutate(Species=str_replace(Species,"Chamaepericlymenum suecicum", "Cornus suecia")) %>%
  #  mutate(Species=str_replace(Species,"Cicerbita alpina", "Lactuca alpina")) %>%
  mutate(Species=str_replace(Species,"Cotoneaster scandinavicus", "Cotoneaster integerrimus")) %>%
  # mutate(Species=str_replace(Species,"Cotoneaster symondsii", "Cotoneaster integrifolius")) %>%
  mutate(Species=str_replace(Species,"Cyanus montanus", "Centaurea montana")) %>%
  #  mutate(Species=str_replace(Species,"Empetrum hermaphroditum", "Empetrum nigrum")) %>%
  mutate(Species=str_replace(Species,"Erysimum virgatum", "Erysimum strictum")) %>%
  #  mutate(Species=str_replace(Species,"Festuca prolifera", "Festuca rubra")) %>%
  mutate(Species=str_replace(Species,"Festuca trachyphylla", "Festuca brevipila")) %>%
  mutate(Species=str_replace(Species,"Galium album", "Galium mollugo")) %>%
  #  mutate(Species=str_replace(Species,"Galium elongatum", "Galium palustre")) %>%
  mutate(Species=str_replace(Species,"Helictotrichon pratense", "Avenula pratensis")) %>%
  mutate(Species=str_replace(Species,"Helictotrichon pubescens", "Avenula pubescens")) %>%
  #  mutate(Species=str_replace(Species,"Hieracium alpina", "Hieracium Alpina")) %>%
  #  mutate(Species=str_replace(Species,"Hieracium alpinum", "Hieracium Alpina")) %>%
  #  mutate(Species=str_replace(Species,"Hieracium hieracium", "Hieracium Hieracium")) %>%
  #  mutate(Species=str_replace(Species,"Hieracium hieracioides", "Hieracium umbellatum")) %>%
  mutate(Species=str_replace(Species,"Hieracium murorum", "Hieracium Hieracium")) %>%
  mutate(Species=str_replace(Species,"Hieracium vulgatum", "Hieracium Vulgata")) %>%
  #  mutate(Species=str_replace(Species,"Hieracium oreadea", "Hieracium Oreadea")) %>%
  #  mutate(Species=str_replace(Species,"Hieracium prenanthoidea", "Hieracium Prenanthoidea")) %>%
  #  mutate(Species=str_replace(Species,"Hieracium vulgata", "Hieracium Vulgata")) %>%
  #  mutate(Species=str_replace(Species,"Hieracium pilosella", "Pilosella officinarum")) %>%
  #  mutate(Species=str_replace(Species,"Hieracium vulgatum", "Hieracium umbellatum")) %>%
  #  mutate(Species=str_replace(Species,"Hierochloã« alpina", "Hierochloë alpina")) %>%
  #  mutate(Species=str_replace(Species,"Hierochloã« hirta", "Hierochloë hirta")) %>%
  #  mutate(Species=str_replace(Species,"Hierochloã« odorata", "Hierochloë odorata")) %>%
  mutate(Species=str_replace(Species,"Hylotelephium maximum", "Sedum telephium")) %>%
  #  mutate(Species=str_replace(Species,"Listera cordata", "Neottia cordata")) %>%
  #  mutate(Species=str_replace(Species,"Leontodon autumnalis", "Scorzoneroides autumnalis")) %>%
  mutate(Species=str_replace(Species,"Lepidotheca suaveolens", "Matricaria discoidea")) %>%
  #  mutate(Species=str_replace(Species,"Loiseleuria procumbens", "Kalmia procumbens")) %>%
  mutate(Species=str_replace(Species,"Malus ×domestica", "Malus domestica")) %>%
  #  mutate(Species=str_replace(Species,"Mycelis muralis", "Lactuca muralis")) %>%
  #  mutate(Species=str_replace(Species,"Omalotheca supina", "Gnaphalium supinum")) %>%
  #  mutate(Species=str_replace(Species,"Omalotheca norvegica", "Gnaphalium norvegicum")) %>%
  #  mutate(Species=str_replace(Species,"Omalotheca sylvatica", "Gnaphalium sylvaticum")) %>%
  #  mutate(Species=str_replace(Species,"Oreopteris limbosperma", "Thelypteris limbosperma")) %>%
  #  mutate(Species=str_replace(Species,"Oxycoccus microcarpus", "Vaccinium microcarpum")) %>%
  #  mutate(Species=str_replace(Species,"Oxycoccus palustris", "Vaccinium oxycoccos")) %>%
  #  mutate(Species=str_replace(Species,"Phalaris minor", "Phalaris arundinacea")) %>%
  mutate(Species=str_replace(Species,"Pilosella peletariana", "Pilosella officinarum")) %>%
  #  mutate(Species=str_replace(Species,"Pinus unicinata", "Pinus mugo")) %>%
  #  mutate(Species=str_replace(Species,"Poa alpigena", "Poa pratensis")) %>%
  mutate(Species=str_replace(Species,"Poa angustifolia", "Poa pratensis")) %>%
  mutate(Species=str_replace(Species,"Poa humilis", "Poa pratensis")) %>%
  #  mutate(Species=str_replace(Species,"Pyrola grandiflora", "Pyrola rotundifolia")) %>%
  mutate(Species=str_replace(Species,"Rosa dumalis", "Rosa vosagiaca")) %>%
  #  mutate(Species=str_replace(Species,"Rumex alpestris", "Rumex acetosa")) %>%
  mutate(Species=str_replace(Species,"Sorbus hybrida", "Hedlundia hybrida")) %>%
  mutate(Species=str_replace(Species,"Spergularia salina", "Spergularia marina")) %>%
  #  mutate(Species=str_replace(Species,"Syringa emodi", "Syringa vulgaris")) %>%
  #  mutate(Species=str_replace(Species,"Taraxacum crocea", "Taraxacum officinale")) %>%
  #  mutate(Species=str_replace(Species,"Taraxacum croceum", "Taraxacum officinale")) %>%
  #  mutate(Species=str_replace(Species,"Trientalis europaea", "Lysimachia europaea")) %>%
  mutate(Species=str_replace(Species,"Trifolium pallidum", "Trifolium pratense"))

# merge species data with indicators
GRUK.species.ind <- merge(x=GRUK.species[,c("Species", "art_dekning", "ParentGlobalID","PolygonID","RuteID")], 
                          y= ind.dat[,c("species","CC", "SS", "RR","Light", "Nitrogen", "Soil_disturbance")],
                          by.x="Species", by.y="species", all.x=T)
summary(GRUK.species.ind)
# checking which species didn't find a match
unique(GRUK.species.ind[is.na(GRUK.species.ind$Light & 
                                is.na(GRUK.species.ind$RR)),'Species'])

## GRUK ruter data handling
names(GRUK.ruter)
# make coordinates numeric
GRUK.ruter <- GRUK.ruter %>% 
  mutate( UTM33_E_ne=as.numeric(UTM33_E_ne) ) %>%
  mutate( UTM33_N_ne=as.numeric(UTM33_N_ne) ) %>%
  mutate( UTM33_E_sw=as.numeric(UTM33_E_sw) ) %>%
  mutate( UTM33_N_sw=as.numeric(UTM33_N_sw) )

# calculate central coordinates for each plot
GRUK.ruter <- GRUK.ruter %>% 
  mutate(UTM33_N = (UTM33_N_ne + UTM33_N_sw)/2) %>%
  mutate(UTM33_E = (UTM33_E_ne + UTM33_E_sw)/2)

# some of the calculations throw NA's because there's only one set of coordinates, coalesce that set into the calculation column 
GRUK.ruter <- GRUK.ruter %>% 
  mutate (UTM33_N = coalesce(UTM33_N,UTM33_N_ne) ) %>%
  mutate (UTM33_E = coalesce(UTM33_E,UTM33_E_ne) )
GRUK.ruter <- GRUK.ruter %>% 
  mutate (UTM33_N = coalesce(UTM33_N,UTM33_N_sw) ) %>%
  mutate (UTM33_E = coalesce(UTM33_E,UTM33_E_sw) )


## GRUK sirkler data handling

## GRUK polygoner data handling

## merge information on mapping units and condition variables from GRUK.sirkler into GRUK.ruter
names(GRUK.ruter)
names(GRUK.sirkler)
GRUK.variables <- merge(x=GRUK.ruter[,c(2,4:11,19:23,25,52:53)], 
                        y=GRUK.sirkler[,c("GlobalID",
                                          "Kartleggingsenhet 1:5000",
                                          "Spor etter slitasje og slitasjebetinget erosjon (%)",
                                          "Dekning % av nakent berg",
                                          "Total dekning % av vedplanter i feltsjikt",
                                          "Dekning % av busker i busksjikt",
                                          "Dekning % av tresjikt",
                                          "Dekning % av problemarter",
                                          "Total dekning % av fremmede arter")], 
                        by.x="GlobalID", by.y="GlobalID", all.x=T)
summary(GRUK.variables)

## merge information on condition and quality from GRUK.polygoner into GRUK.variables
# transform GRUK.variables into spatial object
GRUK.variables <- st_as_sf(GRUK.variables, coords = c("UTM33_E","UTM33_N"),remove=F, crs = 25833)

# transform GRUK.polygoner into spatial object
GRUK.polygoner <- st_as_sf(GRUK.polygoner, wkt = "WKT" ,remove=F, crs = 25833)

tm_shape(GRUK.polygoner) +
  tm_graticules() +
  tm_polygons("PolygonID") +
  tm_shape(GRUK.variables) +
  tm_dots("RuteID")

# run a spatial join to get columns from GRUK.polygoner into GRUK.variables
#GRUK.variables <- st_join(GRUK.variables,GRUK.polygoner[,c(3:4,9,15,18,20,22,24,60)])
#names(GRUK.variables)[1:33]<-c("GlobalID","PolygonID.x","RuteID","RuteID_loknr",
#                               "Dekning_karplanter_feltsjikt","Dekning_moser","Dekning_lav","Dekning_strø",
#                               "Dekning_bar_substrat","Precision","UTM33_E_ne","UTM33_N_ne",
#                               "UTM33_E_sw","UTM33_N_sw","areal(m2)","UTM33_N","UTM33_E","Kartleggingsenhet_1til5000",
#                               "erosjon_prosent","Dekning_nakentberg",
#                               "Totaldekning_vedplanter_feltsjikt","Dekning_busker_busksjikt","Dekning_tresjikt",
#                               "Dekning_problemarter","Totaldekning_fremmedearter","LokalitetID","PolygonID.y",
#                               "Kartleggingsdato","Lokalitetskvalitet","Kommune","Tilstand","Naturmangfold","NiNKartleggingsenheter")

# check how good the spatial join worked
#cbind(GRUK.variables$PolygonID.x,GRUK.variables$PolygonID.y)
#GRUK.variables[7,]
#GRUK.polygoner[GRUK.polygoner$PolygonID=="46-2",]
# some points could not be matched to polygons -> merge by PolygonID instead, drop geometry of GRUK.variables first
GRUK.variables <- st_drop_geometry(GRUK.variables)
names(GRUK.variables)
names(GRUK.polygoner)
GRUK.variables <- merge(x=GRUK.variables, 
                        y=GRUK.polygoner[,c(3:4,9,15,18,20,22,24)], 
                        by.x="PolygonID", by.y="PolygonID", all.x=T)
summary(GRUK.variables) 
summary(as.factor(GRUK.variables$Tilstand)) # no unexpected NA's

# edit the column names
names(GRUK.variables)[1:32]<-c("PolygonID","GlobalID","RuteID","RuteID_loknr",
                         "Dekning_karplanter_feltsjikt","Dekning_moser",
                         "Dekning_lav","Dekning_strø",
                         "Dekning_bar_substrat","Precision",
                         "UTM33_E_ne","UTM33_N_ne","UTM33_E_sw","UTM33_N_sw",
                         "areal_m2","UTM33_N","UTM33_E","Kartleggingsenhet_1til5000",
                         "erosjon_prosent","Dekning_nakentberg",
                         "Totaldekning_vedplanter_feltsjikt","Dekning_busker_busksjikt",
                         "Dekning_tresjikt","Dekning_problemarter","Totaldekning_fremmedearter",
                         "LokalitetID","Aar","Lokalitetskvalitet","Kommune","Tilstand",
                         "Naturmangfold","NiNKartleggingsenheter")



## adding information on ecosystem and condition variables to species+indicator data
names(GRUK.species.ind)
names(GRUK.variables)
GRUK.species.ind <- merge(x=GRUK.species.ind, 
                          y=GRUK.variables[,-c(1,3)], 
                          by.x="ParentGlobalID", by.y="GlobalID", all.x=T)
summary(GRUK.species.ind)
# fixing variable types
GRUK.species.ind <- GRUK.species.ind %>% 
  mutate(Species = as.factor(Species)) %>%
  mutate(areal_m2 = as.numeric(areal_m2)) %>%
  mutate(Kartleggingsenhet_1til5000 = as.factor(Kartleggingsenhet_1til5000)) %>%
  mutate(Dekning_nakentberg = as.numeric(Dekning_nakentberg)) %>%
  mutate(Dekning_problemarter = as.numeric(Dekning_problemarter)) %>%
  mutate(Lokalitetskvalitet = as.factor(Lokalitetskvalitet)) %>%
  mutate(Kommune = as.factor(Kommune)) %>%
  mutate(Tilstand = as.factor(Tilstand)) %>%
  mutate(Naturmangfold = as.factor(Naturmangfold)) %>%
  mutate(NiNKartleggingsenheter = as.factor(NiNKartleggingsenheter))
summary(GRUK.species.ind)

# trimming away the points without information on NiN, species or cover  
GRUK.species.ind <- GRUK.species.ind[!is.na(GRUK.species.ind$Species),]
GRUK.species.ind <- GRUK.species.ind[!is.na(GRUK.species.ind$art_dekning),]
# no NA's for kartleggingsenhet

#rm(GRUK.species)
#rm(GRUK.ruter)


summary(GRUK.species.ind)
head(GRUK.species.ind)


### reference data - data handling

head(natopen_NiN_ref)
head(natopen_NiN_ref_spInfo)

colnames(natopen_NiN_ref)[1] <- "sp"
head(natopen_NiN_ref)

natopen_NiN_ref <- merge(natopen_NiN_ref,natopen_NiN_ref_spInfo[,c(1,4)], by.x="sp", by.y="ScientificName", all.x=T)
unique(natopen_NiN_ref[is.na(natopen_NiN_ref$Phylum),'sp']) # Pucinella does not exist in ind.dat, so we don't care
# we're only interested in vascular plants and ferns, which we have indicators on
unique(natopen_NiN_ref$Phylum)
natopen_NiN_ref <- natopen_NiN_ref %>%
  filter(Phylum %in% c("Magnoliophyta","Pteridophyta"))
unique(natopen_NiN_ref$Phylum)



natopen_NiN_ref$sp
# only genus and species name
natopen_NiN_ref$sp.orig <- natopen_NiN_ref$sp
natopen_NiN_ref$sp <- word(natopen_NiN_ref$sp, 1,2)
natopen_NiN_ref <- natopen_NiN_ref[!is.na(natopen_NiN_ref$sp),]
# merging with indicator values
NiN.natopen <- merge(natopen_NiN_ref,ind.dat[,c(1,3:5,20,23,27)], by.x="sp", by.y="species", all.x=T)
head(NiN.natopen)
summary(NiN.natopen)



NiN.natopen
unique(NiN.natopen$sp)
#NiN.sp$spgr <- as.factor(as.vector(Eco_State$Concept_Data$Species$Species_List$art.code))

# checking which species didn't find a match
unique(NiN.natopen[is.na(NiN.natopen$Light) & is.na(NiN.natopen$Nitrogen) & is.na(NiN.natopen$RR),'sp'])

# fix species name issues
natopen_NiN_ref <- natopen_NiN_ref %>% 
  mutate(sp=str_replace(sp,"Acinos arvensis", "Clinopodium acinos")) %>%
  mutate(sp=str_replace(sp,"Aconitum septentrionale", "Aconitum lycoctonum")) %>%
  mutate(sp=str_replace(sp,"Anagallis minima", "Lysimachia minima")) %>%  
  mutate(sp=str_replace(sp,"Arabis wahlenbergii", "Arabis hirsuta")) %>% 
  mutate(sp=str_replace(sp,"Aristavena setacea", "Deschampsia setacea")) %>% 
  mutate(sp=str_replace(sp,"Atriplex lapponica", "Atriplex longipes")) %>% 
  mutate(sp=str_replace(sp,"Atriplex praecox", "Atriplex longipes")) %>% 
  mutate(sp=str_replace(sp,"Blysmopsis rufa", "Blysmus rufus")) %>% 
  mutate(sp=str_replace(sp,"Carex mackenziei", "Carex norvegica")) %>% 
  mutate(sp=str_replace(sp,"Carex xvacillans", "Carex vacillans")) %>% 
  mutate(sp=str_replace(sp,"Cicerbita alpina", "Lactuca alpina")) %>% 
  mutate(sp=str_replace(sp,"Cirsium acaulon", "Cirsium acaule")) %>%
  mutate(sp=str_replace(sp,"Cotoneaster scandinavicus", "Cotoneaster integerrimus")) %>%
  mutate(sp=str_replace(sp,"Dactylorhiza viridis", "Coeloglossum viride")) %>%
  mutate(sp=str_replace(sp,"Elymus alaskanus", "Elymus kronokensis")) %>%
  mutate(sp=str_replace(sp,"Elymus repens", "Elytrigia repens")) %>%
  mutate(sp=str_replace(sp,"Epilobium angustifolium", "Chamaenerion angustifolium")) %>%
  mutate(sp=str_replace(sp,"Erysimum virgatum", "Erysimum strictum")) %>%
  mutate(sp=str_replace(sp,"Euphrasia frigida", "Euphrasia wettsteinii")) %>%
  mutate(sp=str_replace(sp,"Hedlundia meinichii", "Sorbus aucuparia")) %>%
  mutate(sp=str_replace(sp,"Hylotelephium maximum", "Hylotelephium telephium")) %>%
  mutate(sp=str_replace(sp,"Juncus arcuatus", "Luzula arcuata")) %>%
  mutate(sp=str_replace(sp,"Kali turgida", "Kali turgidum")) %>%
  mutate(sp=str_replace(sp,"Kobresia myosuroides", "Carex myosuroides")) %>%
  mutate(sp=str_replace(sp,"Luzula pallidula", "Luzula pallescens")) %>%
  mutate(sp=str_replace(sp,"Lychnis alpina", "Viscaria alpina")) %>%
  mutate(sp=str_replace(sp,"Minuartia rubella", "Sabulina rubella")) %>%
  mutate(sp=str_replace(sp,"Minuartia stricta", "Sabulina stricta")) %>%
  mutate(sp=str_replace(sp,"Ononis arvensis", "Ononis spinosa")) %>%
  mutate(sp=str_replace(sp,"Papaver dahlianum", "Papaver radicatum")) %>%
  mutate(sp=str_replace(sp,"Papaver lapponicum", "Papaver radicatum")) %>%
  mutate(sp=str_replace(sp,"Pentanema salicinum", "Inula salicina")) %>%
  mutate(sp=str_replace(sp,"Poa alpigena", "Poa pratensis")) %>%
  mutate(sp=str_replace(sp,"Poa humilis", "Poa pratensis")) %>%
  mutate(sp=str_replace(sp,"Polygonum boreale", "Polygonum aviculare")) %>%
  mutate(sp=str_replace(sp,"Polygonum excelsius", "Polygonum aviculare")) %>%
  mutate(sp=str_replace(sp,"Polygonum norvegicum", "Polygonum oxyspermum")) %>%
  mutate(sp=str_replace(sp,"Potentilla tabernaemontani", "Potentilla verna")) %>%
  mutate(sp=str_replace(sp,"Puccinellia retroflexa", "Puccinellia capillaris")) %>%
  mutate(sp=str_replace(sp,"Pyrola norvegica", "Pyrola rotundifolia")) %>%
  mutate(sp=str_replace(sp,"Roegneria canina", "Elymus caninus")) %>%
  mutate(sp=str_replace(sp,"Rumex graminifolius", "Rumex acetosella")) %>%
  mutate(sp=str_replace(sp,"Salicornia pojarkovae", "Salicornia procumbens")) %>%
  mutate(sp=str_replace(sp,"Schedonorus elatior", "Schedonorus arundinaceus")) %>%
  mutate(sp=str_replace(sp,"Selinum carvifolium", "Selinum carvifolia")) %>%
  mutate(sp=str_replace(sp,"Spergula marina", "Spergularia marina")) %>%
  mutate(sp=str_replace(sp,"Spergularia salina", "Spergularia marina")) %>%
  mutate(sp=str_replace(sp,"Thinopyrum junceiforme", "Elytrigia juncea"))

#ind.dat[2556,'species'] <- "Saxifraga osloënsis"
#ind.dat[17,'species'] <- "Hierochloë odorata"
#ind.dat[9,'species'] <- "Hippophaë rhamnoides"


# merging with indicator values
NiN.natopen <- merge(natopen_NiN_ref,ind.dat[,c(1,3:5,20,23,27)], by.x="sp", by.y="species", all.x=T)
# checking which species didn't find a match
unique(NiN.natopen[is.na(NiN.natopen$Light) & is.na(NiN.natopen$Nitrogen) & is.na(NiN.natopen$RR),'sp'])
# ok now


# translating the abundance classes into %-cover
coverscale <- data.frame(orig=0:6,
                         cov=c(0, 1/32 ,1/8, 3/8, 0.6, 4/5, 1)
)

NiN.natopen.cov <- NiN.natopen
colnames(NiN.natopen.cov)
for (i in 2:71) {
  NiN.natopen.cov[,i] <- coverscale[,2][ match(NiN.natopen[,i], 0:6 ) ]
}

summary(NiN.natopen)
summary(NiN.natopen.cov)
```
leaving us with the monitoring data including plant indicators (GRUK.species.ind) and the reference data including plant indicators (NiN.natopen.cov)

```{r}
head(GRUK.species.ind)
head(NiN.natopen.cov)
```
For each ecosystem type with a NiN species list, we can calculate a community weighted mean (CWM) for the relevant functional plant indicators.
For semi-natural ecosystems, we are testing "Light", "Moisture", "Soil_reaction_pH", "Nitrogen", "Phosphorus", "Grazing_mowing", and "Soil_disturbance". In order to get distributions of CWMs rather than one single value (for comparison with the empirical testing data), the NiN lists can be bootstrapped.

##### bootstrap function for frequency abundance
- function to calculate community weighted means of selected indicator values (ind)
- for species lists (sp) with given abundances in percent (or on a scale from 0 to 1) in one or more 'sites' (abun)
- with a given number of iterations (iter),
- with species given a certain minimum abundance occurring in all bootstraps (obl), and
- with a given re-sampling ratio of the original species list (rat)
- in every bootstrap iteration the abundance of the sampled species can be randomly changed by a limited amount if wished by introducing a re-sampling of abundance values from adjacent abundance steps with a certain probability (var.abun)

```{r, echo = F}
indBoot.freq <- function(sp,abun,ind,iter,obl,rat=2/3,var.abun=F) {
  
  ind.b <- matrix(nrow=iter,ncol=length(colnames(abun)))
  colnames(ind.b) <- colnames(abun)
  ind.b <- as.data.frame(ind.b)  
  
  ind <- as.data.frame(ind)
  ind.list <- as.list(1:length(colnames(ind)))
  names(ind.list) <- colnames(ind)
  
  for (k in 1:length(colnames(ind)) ) {
    ind.list[[k]] <- ind.b }
  
  for (j in 1:length(colnames(abun)) ) {
    
    dat <- cbind(sp,abun[,j],ind)
    dat <- dat[dat[,2]>0,]            # only species that are present in the ecosystem
    dat <- dat[!is.na(dat[,3]),]      # only species that have indicator values
    
    for (i in 1:iter) {
      
      speciesSample <- sample(dat$sp[dat[,2] < obl], size=round( (length(dat$sp)-length(dat$sp[dat[,2]>=obl])) *rat,0), replace=F)  
      dat.b <- rbind(dat[dat[,2] >= obl,],
                     dat[match(speciesSample,dat$sp),]
      )
      
      if (var.abun==T) {
        for (m in 1:nrow(coverscale[-1,]) ) {
          xxx <- dat.b[dat.b[,2]==coverscale[-1,][m,2],2]
          if ( m==1 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.5, 0.5, 0.0, 0.0, 0.0, 0.0, 0.0) ,size=length(xxx),replace=T) }
          if ( m==2 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.2, 0.3, 0.5, 0.0, 0.0, 0.0, 0.0) ,size=length(xxx),replace=T) }
          if ( m==3 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.0, 0.2, 0.3, 0.5, 0.0, 0.0, 0.0) ,size=length(xxx),replace=T) }
          if ( m==4 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.0, 0.0, 0.2, 0.3, 0.5, 0.0, 0.0) ,size=length(xxx),replace=T) }
          if ( m==5 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.0, 0.0, 0.0, 0.2, 0.3, 0.5, 0.0) ,size=length(xxx),replace=T) }
          if ( m==6 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.0, 0.0, 0.0, 0.0, 0.2, 0.3, 0.5) ,size=length(xxx),replace=T) }
        }
        dat.b[!is.na(dat.b[,2]) & dat.b[,2]<=(0),2] <- 0.01
        dat.b[!is.na(dat.b[,2]) & dat.b[,2]>1,2] <- 1
      }
      
      for (k in 1:length(colnames(ind))) {
        
        if ( nrow(dat.b)>2 ) {
          
          ind.b <- sum(dat.b[!is.na(dat.b[,2+k]),2] * dat.b[!is.na(dat.b[,2+k]),2+k] , na.rm=T) / sum(dat.b[!is.na(dat.b[,2+k]),2],na.rm=T)
          ind.list[[k]][i,j] <- ind.b
          
        } else {ind.list[[k]][i,j] <- NA}
        
      }
 
#      print(paste(i,"",j)) 
    }
    
  }
  return(ind.list)
}
```

Running the bootstraps

```{r, echo = T, eval = F}
colnames(NiN.natopen)
# 1st column is the species
# 6th-71st column is the abundances of sp in different ecosystem types
# 74th-79th column is the indicator values of the respective species
# we choose 1000 iterations
# species with abundance 1 (i.e. a max of 100%, must be included in each sample)
# each sample re-samples 1/3 of the number of species
# the abundance of the re-sampled species may vary (see bootstrap function for details)
natopen.ref.cov <- indBoot.freq(sp=NiN.natopen.cov[,1],abun=NiN.natopen.cov[,6:71],ind=NiN.natopen.cov[,74:79],iter=1000,obl=1,rat=1/3,var.abun=T)

### fixing NaN's
for (i in 1:length(natopen.ref.cov) ) {
  for (j in 1:ncol(natopen.ref.cov[[i]]) ) {
    v <- natopen.ref.cov[[i]][,j]
    v[is.nan(v)] <- NA
    natopen.ref.cov[[i]][,j] <- v
  }
}
```

```{r, include = F}
# Data from cache
natopen.ref.cov<-readRDS(paste0(here::here(), "/data/cache/natopen.ref.cov.RDS"))
```

```{r}
head(natopen.ref.cov[[1]])
```

This results in an R-list with a slot for every selected indicator, and in every slot there's a data frame with as many columns as there are NiN species lists and as many rows as there were iterations in the bootstrap.
Next, we need to derive scaling values from these bootstrap-lists (the columns) for every mapping unit in NiN. Here, we define things in the following way:

- Median = reference values
- 0.025 and 0.975 quantiles = lower and upper limit values
- min and max of the respective indicator's scale = min/max values

```{r, echo = F}

# NiN-types where each type is represented by one species list (including when one species list represents two NiN-types)
names(natopen.ref.cov[["Light"]])
x <- 1:66

# checking the actual NiN-types in the natopen lists
natopen.NiNtypes <- colnames(natopen.ref.cov[["Light"]])
natopen.NiNtypes[-x] <- substr(natopen.NiNtypes[-x], 1, nchar(natopen.NiNtypes[-x])-1)
natopen.NiNtypes

# 7 indicator-value indicators: Grimes C, S and R, Tyler's Light, Nitrogen, and Soil_disturbance
indEll.n=6
# creating a table to hold:
# Tyler: the 0.5 quantile (median), 0.05 quantile and  0.95 quantile for each NiN-type
# for every nature type (nrows)
tab <- matrix(ncol=3*indEll.n, nrow=length(unique(natopen.NiNtypes)) ) # 66 basic ecosystem types
# coercing the values into the table


for (i in 1:length(x) ) {
  tab[i,1:3] <- quantile(as.matrix(natopen.ref.cov[["CC"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
  tab[i,4:6] <- quantile(as.matrix(natopen.ref.cov[["SS"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
  tab[i,7:9] <- quantile(as.matrix(natopen.ref.cov[["RR"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
  tab[i,10:12] <- quantile(as.matrix(natopen.ref.cov[["Light"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
  tab[i,13:15] <- quantile(as.matrix(natopen.ref.cov[["Nitrogen"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
  tab[i,16:18] <- quantile(as.matrix(natopen.ref.cov[["Soil_disturbance"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
  
}

tab <- as.data.frame(tab)
tab$NiN <- NA
tab$NiN[1:length(x)] <- names(natopen.ref.cov[[1]])[x]
tab


# making it a proper data frame
dim(tab)
round(tab[,1:18],digits=2)

colnames(tab) <- c("CC_q2.5","CC_q50","CC_q97.5",
                   "SS_q2.5","SS_q50","SS_q97.5",
                   "RR_q2.5","RR_q50","RR_q97.5",
                   "Light_q2.5","Light_q50","Light_q97.5",
                   "Nitrogen_q2.5","Nitrogen_q50","Nitrogen_q97.5",
                   "Soil_disturbance_q2.5","Soil_disturbance_q50","Soil_disturbance_q97.5",
                   "NiN")
summary(tab)
tab$NiN
#tab$NiN <- gsub("C", "C-", tab$NiN) # add extra hyphen after C for NiN-types
tab


# restructuring into separate indicators for lower (q2.5) and higher (q97.5) than reference value (=median, q50)
y.Light <- numeric(length=nrow(tab)*2)
y.Light[((1:dim(tab)[1])*2)-1] <- tab$Light_q2.5 
y.Light[((1:dim(tab)[1])*2)] <- tab$Light_q97.5 

y.CC <- numeric(length=nrow(tab)*2)
y.CC[((1:dim(tab)[1])*2)-1] <- tab$CC_q2.5 
y.CC[((1:dim(tab)[1])*2)] <- tab$CC_q97.5 

y.SS <- numeric(length=nrow(tab)*2)
y.SS[((1:dim(tab)[1])*2)-1] <- tab$SS_q2.5 
y.SS[((1:dim(tab)[1])*2)] <- tab$SS_q97.5 

y.Nitrogen <- numeric(length=nrow(tab)*2)
y.Nitrogen[((1:dim(tab)[1])*2)-1] <- tab$Nitrogen_q2.5 
y.Nitrogen[((1:dim(tab)[1])*2)] <- tab$Nitrogen_q97.5 

y.RR <- numeric(length=nrow(tab)*2)
y.RR[((1:dim(tab)[1])*2)-1] <- tab$RR_q2.5 
y.RR[((1:dim(tab)[1])*2)] <- tab$RR_q97.5 

y.Soil_disturbance <- numeric(length=nrow(tab)*2)
y.Soil_disturbance[((1:dim(tab)[1])*2)-1] <- tab$Soil_disturbance_q2.5 
y.Soil_disturbance[((1:dim(tab)[1])*2)] <- tab$Soil_disturbance_q97.5 

# creating final objects holding the reference and limit values for all indicators

# ref object for indicators
natopen.ref.cov.val <- data.frame(N1=rep('natopen',(nrow(tab)*2*indEll.n)),
                                  hoved=c(rep('NA',(nrow(tab)*2*indEll.n))),
                                  grunn=c(rep(rep(tab$NiN,each=2),indEll.n)),
                                  county=rep('all',(nrow(tab)*2*indEll.n)),
                                  region=rep('all',(nrow(tab)*2*indEll.n)),
                                  Ind=c(rep(c('CC1','CC2'),nrow(tab)),
                                        rep(c('SS1','SS2'),nrow(tab)),
                                        rep(c('RR1','RR2'),nrow(tab)),
                                        rep(c('Light1','Light2'),nrow(tab)),
                                        rep(c('Nitrogen1','Nitrogen2'),nrow(tab)),
                                        rep(c('Soil_disturbance1','Soil_disturbance2'),nrow(tab))
                                  ),
                                  Rv=c(rep(tab$CC_q50,each=2),
                                       rep(tab$SS_q50,each=2),
                                       rep(tab$RR_q50,each=2),
                                       rep(tab$Light_q50,each=2),
                                       rep(tab$Nitrogen_q50,each=2),
                                       rep(tab$Soil_disturbance_q50,each=2)
                                  ),
                                  Gv=c(y.CC,y.SS,y.RR,y.Light,y.Nitrogen,y.Soil_disturbance),
                                  maxmin=c(rep(c(0,1),nrow(tab)), # CC
                                           rep(c(0,1),nrow(tab)), # SS
                                           rep(c(0,1),nrow(tab)), # RR
                                           rep(c(1,7),nrow(tab)),  # 7 levels of Light
                                           rep(c(1,9),nrow(tab)),  # 9 levels of Nitrogen
                                           rep(c(1,9),nrow(tab))  # 9 levels of Soil_disturbance
                                  )
)

natopen.ref.cov.val
natopen.ref.cov.val$grunn <- as.factor(natopen.ref.cov.val$grunn)
natopen.ref.cov.val$Ind <- as.factor(natopen.ref.cov.val$Ind)
summary(natopen.ref.cov.val)
head(natopen.ref.cov.val)

### add scaling values for cover of slitasje, alien sp., problem sp., woody species
natopen.ref.cov.val <- natopen.ref.cov.val %>% add_row(N1="natopen",
                                hoved=rep("T2",6),
                                grunn=rep(c("T2-C-7","T2-C-8"),3),
                                county=rep("all",6),
                                region=rep("all",6),
                                Ind=c("aliens","aliens","erosion","erosion","shrub","shrub"),
                                Rv=c(100,100,100,100,90,90),
                                Gv=c(95,95,93.75,93.75,75,75),
                                maxmin=c(0,0,0,0,25,25)
                )


```

```{r}
head(natopen.ref.cov.val)
```

Once test data (ANO, GRUK) and the scaling values from the reference data are in place, we can calculate community-weighted means (CWM) of the selected indicators for the ANO and GRUK community data and scale them against the scaling values from the reference distribution. Note that we scale each ANO/GRUK plot's CWM against either the lower threshold value and the min value OR the upper threshold value and the max value based on whether the CWM is smaller or higher than the reference value. Since the scaled values for both sides range between 0 and 1, we generate separate lower and upper condition indicators for each  functional plant indicator. An ANO/GRUK plot can only have a scaled value in either the lower or the upper indicator (the other one will be 'NA'), except for the unlikely event that the CWM exactly matches the reference value, in which case both lower and upper indicator will receive a scaled indicator value of 1.

Here is the scaling function
```{r}

#### scaled values ####
r.s <- 1    # reference value
l.s <- 0.6  # limit value
a.s <- 0    # abscence of indicator, or indicator at maximum

#### function for calculating scaled values for measured value ####

## scaling function including truncation
scal <- function() {
  # place to hold the result
   x <- numeric()
  if (maxmin < ref) {
    # values >= the reference value equal 1
    if (val >= ref) {x <- 1}
    # values < the reference value and >= the limit value can be deducted from the linear relationship between these two
    if (val < ref & val >= lim) {x <- (l.s + (val-lim) * ( (r.s-l.s) / (ref-lim) ) )}
    # values < the limit value and > maxmin can be deducted from the linear relationship between these two
    if (val < lim & val > maxmin) {x <- (a.s + (val-maxmin) * ( (l.s-a.s) / (lim-maxmin) ) )}
    # value equals or lower than maxmin
    if (val <= maxmin) {x <-0}
  } else {
    # values <= the reference value equal 1
    if (val <= ref) {x <- 1}
    # values > the reference value and <= the limit value can be deducted from the linear relationship between these two
    if (val > ref & val <= lim) {x <- ( r.s - ( (r.s - l.s) * (val - ref) / (lim - ref) ) )}
    # values > the limit value and < maxmin can be deducted from the linear relationship between these two
    if (val > lim) {x <- ( l.s - (l.s * (val - lim) / (maxmin - lim) ) )}
    # value equals or larger than maxmin
    if (val >= maxmin) {x <-0}
  }
  return(x)
  
}

## scaling function without truncation
scal.2 <- function() {
  # place to hold the result
  x <- numeric()
  if (maxmin < ref) {
    # values >= the reference value estimated from the linear relationship for lim < x < ref (line below)
    if (val >= ref) {x <- (l.s + (val-lim) * ( (r.s-l.s) / (ref-lim) ) )}
    # values < the reference value and >= the limit value can be deducted from the linear relationship between these two
    if (val < ref & val >= lim) {x <- (l.s + (val-lim) * ( (r.s-l.s) / (ref-lim) ) )}
    # values < the limit value and > maxmin can be deducted from the linear relationship between these two
    if (val < lim & val > maxmin) {x <- (a.s + (val-maxmin) * ( (l.s-a.s) / (lim-maxmin) ) )}
    # value equal or lower than maxmin
    if (val <= maxmin) {x <-0}
  } else {
    # values <= the reference value estimated from the linear relationship for lim < x < ref (line below)
    if (val <= ref) {x <- ( r.s - ( (r.s - l.s) * (val - ref) / (lim - ref) ) )}
    # values > the reference value and <= the limit value can be deducted from the linear relationship between these two
    if (val > ref & val <= lim) {x <- ( r.s - ( (r.s - l.s) * (val - ref) / (lim - ref) ) )}
    # values > the limit value and < maxmin can be deducted from the linear relationship between these two
    if (val > lim & val < maxmin) {x <- ( l.s - (l.s * (val - lim) / (maxmin - lim) ) )}
    # value equal og larger than maxmin
    if (val >= maxmin) {x <-0}
  }
  return(x)
  
}

```

We then can prepare a list of data frames to hold the results and perform the scaling according to the principles described in NINA report 1967 (Töpper and Jakobsson 2021)
```{r, include=FALSE}

colnames(GRUK.variables)
levels(as.factor(GRUK.variables$Kartleggingsenhet_1til5000)) # NiN types in data
levels(natopen.ref.cov.val$grunn) # NiN types in reference
#### creating dataframe to hold the results for natopens ####
# all GRUK points
nrow(GRUK.variables)
GRUK.natopen <- GRUK.variables

head(GRUK.natopen)
# update row-numbers
row.names(GRUK.natopen) <- 1:nrow(GRUK.natopen)
head(GRUK.natopen)
dim(GRUK.natopen)
colnames(GRUK.natopen)

length(levels(as.factor(GRUK.natopen$PolygonID)))
length(levels(as.factor(GRUK.natopen$RuteID)))
summary(as.factor(GRUK.natopen$RuteID))
# none are double

unique(GRUK.natopen$Kartleggingsenhet_1til5000)
GRUK.natopen$Kartleggingsenhet_1til5000 <- as.factor(GRUK.natopen$Kartleggingsenhet_1til5000)
summary(GRUK.natopen$Kartleggingsenhet_1til5000)

results.natopen.GRUK <- list()
ind <- unique(natopen.ref.cov.val$Ind)
# choose columns for site description
colnames(GRUK.natopen)
results.natopen.GRUK[['original']] <- GRUK.natopen
# drop geometry
#st_geometry(results.natopen.GRUK[['original']]) <- NULL
results.natopen.GRUK[['original']] <- as.data.frame(results.natopen.GRUK[['original']])

# add columns for indicators
nvar.site <- ncol(results.natopen.GRUK[['original']])
for (i in 1:length(ind) ) {results.natopen.GRUK[['original']][,i+nvar.site] <- NA}
colnames(results.natopen.GRUK[['original']])[(nvar.site+1):(length(ind)+nvar.site)] <- paste(ind)
for (i in (nvar.site+1):(length(ind)+nvar.site) ) {results.natopen.GRUK[['original']][,i] <- as.numeric(results.natopen.GRUK[['original']][,i])}
summary(results.natopen.GRUK[['original']])
results.natopen.GRUK[['original']]$Kommune <- as.factor(results.natopen.GRUK[['original']]$Kommune)
results.natopen.GRUK[['original']]$GlobalID <- as.factor(results.natopen.GRUK[['original']]$GlobalID)
results.natopen.GRUK[['original']]$PolygonID <- as.factor(results.natopen.GRUK[['original']]$PolygonID)
results.natopen.GRUK[['original']]$uRute_ID <- as.factor(results.natopen.GRUK[['original']]$RuteID)
results.natopen.GRUK[['original']]$Kartleggingsenhet_1til5000 <- as.factor(results.natopen.GRUK[['original']]$Kartleggingsenhet_1til5000)

# roll out
results.natopen.GRUK[['scaled']] <- results.natopen.GRUK[['non-truncated']] <- results.natopen.GRUK[['original']]


```

```{r, echo = T, eval = F, include = F}

#### calculating scaled and non-truncated values for the indicators based on the dataset ####
for (i in 1:nrow(GRUK.natopen) ) {  #
  tryCatch({
    print(i)
    print(paste(GRUK.natopen$PolygonID[i]))
    print(paste(GRUK.natopen$RuteID[i]))
    #    GRUK.natopen$Hovedoekosystem_sirkel[i]
    #    GRUK.natopen$Hovedoekosystem_rute[i]
    
    
    
    # if the GRUK.hovedtype exists in the reference
    #    if (GRUK.natopen$hovedtype_rute[i] %in% unique(substr(natopen.ref.cov.val$grunn,1,2)) ) {
    
    # if there is any species present in current GRUK point  
    if ( length(GRUK.species.ind[GRUK.species.ind$ParentGlobalID==as.character(GRUK.natopen$GlobalID[i]),'Species']) > 0 ) {
      
      
      # Grime's C
      
      dat <- GRUK.species.ind[GRUK.species.ind$ParentGlobalID==as.character(GRUK.natopen$GlobalID[i]),c('art_dekning','CC')]
      results.natopen.GRUK[['original']][i,'richness'] <- nrow(dat)
      dat <- dat[!is.na(dat$CC),]
      
      if ( nrow(dat)>0 ) {
        
        val <- sum(dat[,'art_dekning'] * dat[,'CC'],na.rm=T) / sum(dat[,'art_dekning'],na.rm=T)
        # lower part of distribution
        ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='CC1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),"_BN",sep=""),'Rv']
        lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='CC1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),"_BN",sep=""),'Gv']
        maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='CC1' & natopen.ref.cov.val$grunn==as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),'maxmin']
        # coercing x into results.natopen.GRUK dataframe
        results.natopen.GRUK[['scaled']][i,'CC1'] <- scal() 
        results.natopen.GRUK[['non-truncated']][i,'CC1'] <- scal.2() 
        results.natopen.GRUK[['original']][i,'CC1'] <- val 
        
        # upper part of distribution
        ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='CC2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),"_BN",sep=""),'Rv']
        lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='CC2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),"_BN",sep=""),'Gv']
        maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='CC2' & natopen.ref.cov.val$grunn==as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),'maxmin']
        # coercing x into results.natopen.GRUK dataframe
        results.natopen.GRUK[['scaled']][i,'CC2'] <- scal() 
        results.natopen.GRUK[['non-truncated']][i,'CC2'] <- scal.2() 
        results.natopen.GRUK[['original']][i,'CC2'] <- val
        
      }
      
      
      # Grime's S
      dat <- GRUK.species.ind[GRUK.species.ind$ParentGlobalID==as.character(GRUK.natopen$GlobalID[i]),c('art_dekning','SS')]
      results.natopen.GRUK[['original']][i,'richness'] <- nrow(dat)
      dat <- dat[!is.na(dat$SS),]
      
      if ( nrow(dat)>0 ) {
        
        val <- sum(dat[,'art_dekning'] * dat[,'SS'],na.rm=T) / sum(dat[,'art_dekning'],na.rm=T)
        # lower part of distribution
        ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='SS1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),"_BN",sep=""),'Rv']
        lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='SS1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),"_BN",sep=""),'Gv']
        maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='SS1' & natopen.ref.cov.val$grunn==as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),'maxmin']
        # coercing x into results.natopen.GRUK dataframe
        results.natopen.GRUK[['scaled']][i,'SS1'] <- scal() 
        results.natopen.GRUK[['non-truncated']][i,'SS1'] <- scal.2() 
        results.natopen.GRUK[['original']][i,'SS1'] <- val
        
        # upper part of distribution
        ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='SS2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),"_BN",sep=""),'Rv']
        lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='SS2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),"_BN",sep=""),'Gv']
        maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='SS2' & natopen.ref.cov.val$grunn==as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),'maxmin']
        # coercing x into results.natopen.GRUK dataframe
        results.natopen.GRUK[['scaled']][i,'SS2'] <- scal() 
        results.natopen.GRUK[['non-truncated']][i,'SS2'] <- scal.2() 
        results.natopen.GRUK[['original']][i,'SS2'] <- val
        
      }
      
      
      # Grime's R
      dat <- GRUK.species.ind[GRUK.species.ind$ParentGlobalID==as.character(GRUK.natopen$GlobalID[i]),c('art_dekning','RR')]
      results.natopen.GRUK[['original']][i,'richness'] <- nrow(dat)
      dat <- dat[!is.na(dat$RR),]
      
      if ( nrow(dat)>0 ) {
        
        val <- sum(dat[,'art_dekning'] * dat[,'RR'],na.rm=T) / sum(dat[,'art_dekning'],na.rm=T)
        # lower part of distribution
        ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='RR1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),"_BN",sep=""),'Rv']
        lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='RR1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),"_BN",sep=""),'Gv']
        maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='RR1' & natopen.ref.cov.val$grunn==as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),'maxmin']
        # coercing x into results.natopen.GRUK dataframe
        results.natopen.GRUK[['scaled']][i,'RR1'] <- scal() 
        results.natopen.GRUK[['non-truncated']][i,'RR1'] <- scal.2() 
        results.natopen.GRUK[['original']][i,'RR1'] <- val
        
        # upper part of distribution
        ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='RR2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),"_BN",sep=""),'Rv']
        lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='RR2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),"_BN",sep=""),'Gv']
        maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='RR2' & natopen.ref.cov.val$grunn==as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),'maxmin']
        # coercing x into results.natopen.GRUK dataframe
        results.natopen.GRUK[['scaled']][i,'RR2'] <- scal() 
        results.natopen.GRUK[['non-truncated']][i,'RR2'] <- scal.2() 
        results.natopen.GRUK[['original']][i,'RR2'] <- val
        
      }
      
      
      # Light
      dat <- GRUK.species.ind[GRUK.species.ind$ParentGlobalID==as.character(GRUK.natopen$GlobalID[i]),c('art_dekning','Light')]
      results.natopen.GRUK[['original']][i,'richness'] <- nrow(dat)
      dat <- dat[!is.na(dat$Light),]
      
      if ( nrow(dat)>0 ) {
        
        val <- sum(dat[,'art_dekning'] * dat[,'Light'],na.rm=T) / sum(dat[,'art_dekning'],na.rm=T)
        # lower part of distribution
        ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Light1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),"_BN",sep=""),'Rv']
        lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Light1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),"_BN",sep=""),'Gv']
        maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Light1' & natopen.ref.cov.val$grunn==as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),'maxmin']
        # coercing x into results.natopen.GRUK dataframe
        results.natopen.GRUK[['scaled']][i,'Light1'] <- scal() 
        results.natopen.GRUK[['non-truncated']][i,'Light1'] <- scal.2() 
        results.natopen.GRUK[['original']][i,'Light1'] <- val
        
        # upper part of distribution
        ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Light2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),"_BN",sep=""),'Rv']
        lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Light2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),"_BN",sep=""),'Gv']
        maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Light2' & natopen.ref.cov.val$grunn==as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),'maxmin']
        # coercing x into results.natopen.GRUK dataframe
        results.natopen.GRUK[['scaled']][i,'Light2'] <- scal() 
        results.natopen.GRUK[['non-truncated']][i,'Light2'] <- scal.2() 
        results.natopen.GRUK[['original']][i,'Light2'] <- val
        
        
      }
      
      
      
      
      # Nitrogen
      dat <- GRUK.species.ind[GRUK.species.ind$ParentGlobalID==as.character(GRUK.natopen$GlobalID[i]),c('art_dekning','Nitrogen')]
      results.natopen.GRUK[['original']][i,'richness'] <- nrow(dat)
      dat <- dat[!is.na(dat$Nitrogen),]
      
      if ( nrow(dat)>0 ) {
        
        val <- sum(dat[,'art_dekning'] * dat[,'Nitrogen'],na.rm=T) / sum(dat[,'art_dekning'],na.rm=T)
        # lower part of distribution
        ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Nitrogen1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),"_BN",sep=""),'Rv']
        lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Nitrogen1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),"_BN",sep=""),'Gv']
        maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Nitrogen1' & natopen.ref.cov.val$grunn==as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),'maxmin']
        # coercing x into results.natopen.GRUK dataframe
        results.natopen.GRUK[['scaled']][i,'Nitrogen1'] <- scal() 
        results.natopen.GRUK[['non-truncated']][i,'Nitrogen1'] <- scal.2() 
        results.natopen.GRUK[['original']][i,'Nitrogen1'] <- val
        
        
        # upper part of distribution
        ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Nitrogen2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),"_BN",sep=""),'Rv']
        lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Nitrogen2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),"_BN",sep=""),'Gv']
        maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Nitrogen2' & natopen.ref.cov.val$grunn==as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),'maxmin']
        # coercing x into results.natopen.GRUK dataframe
        results.natopen.GRUK[['scaled']][i,'Nitrogen2'] <- scal() 
        results.natopen.GRUK[['non-truncated']][i,'Nitrogen2'] <- scal.2() 
        results.natopen.GRUK[['original']][i,'Nitrogen2'] <- val
        
      }
      
      
      
      
      
      
      # Soil_disturbance
      dat <- GRUK.species.ind[GRUK.species.ind$ParentGlobalID==as.character(GRUK.natopen$GlobalID[i]),c('art_dekning','Soil_disturbance')]
      results.natopen.GRUK[['original']][i,'richness'] <- nrow(dat)
      dat <- dat[!is.na(dat$Soil_disturbance),]
      
      if ( nrow(dat)>0 ) {
        
        val <- sum(dat[,'art_dekning'] * dat[,'Soil_disturbance'],na.rm=T) / sum(dat[,'art_dekning'],na.rm=T)
        # lower part of distribution
        ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Soil_disturbance1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),"_BN",sep=""),'Rv']
        lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Soil_disturbance1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),"_BN",sep=""),'Gv']
        maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Soil_disturbance1' & natopen.ref.cov.val$grunn==as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),'maxmin']
        # coercing x into results.natopen.GRUK dataframe
        results.natopen.GRUK[['scaled']][i,'Soil_disturbance1'] <- scal() 
        results.natopen.GRUK[['non-truncated']][i,'Soil_disturbance1'] <- scal.2() 
        results.natopen.GRUK[['original']][i,'Soil_disturbance1'] <- val
        
        
        # upper part of distribution
        ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Soil_disturbance2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),"_BN",sep=""),'Rv']
        lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Soil_disturbance2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),"_BN",sep=""),'Gv']
        maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Soil_disturbance2' & natopen.ref.cov.val$grunn==as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),'maxmin']
        # coercing x into results.natopen.GRUK dataframe
        results.natopen.GRUK[['scaled']][i,'Soil_disturbance2'] <- scal() 
        results.natopen.GRUK[['non-truncated']][i,'Soil_disturbance2'] <- scal.2() 
        results.natopen.GRUK[['original']][i,'Soil_disturbance2'] <- val
        
      }
      
    }
    #    }
    
       # Area cover without alien species
      dat <- GRUK.natopen[i,"Totaldekning_fremmedearter"]
      
      if ( !is.na(dat) ) {
        
        val <- 100-dat
        # one-sided indicator
        ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='aliens' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"])),'Rv']
        lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='aliens' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"])),'Gv']
        maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='aliens' & natopen.ref.cov.val$grunn==as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),'maxmin']
        # coercing x into results.natopen.GRUK dataframe
        results.natopen.GRUK[['scaled']][i,'aliens'] <- scal() 
        results.natopen.GRUK[['non-truncated']][i,'aliens'] <- scal() 
        results.natopen.GRUK[['original']][i,'aliens'] <- val
        
      }    
      
      ## run the indicators that are not dependent on species in the plot
      # Area cover without erosion
      dat <- GRUK.natopen[i,"erosjon_prosent"]
      
      if ( !is.na(dat) ) {
        
        val <- 100-dat
        # one-sided indicator
        ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='erosion' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"])),'Rv']
        lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='erosion' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"])),'Gv']
        maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='erosion' & natopen.ref.cov.val$grunn==as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),'maxmin']
        # coercing x into results.natopen.GRUK dataframe
        results.natopen.GRUK[['scaled']][i,'erosion'] <- scal() 
        results.natopen.GRUK[['non-truncated']][i,'erosion'] <- scal() 
        results.natopen.GRUK[['original']][i,'erosion'] <- val
        
      } 
      
      # area cover without shrub cover
      dat <- GRUK.natopen[i,"Dekning_busker_busksjikt"]
      
      if ( !is.na(dat) ) {
        
        val <- 100-dat
        # one-sided indicator
        ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='shrub' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"])),'Rv']
        lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='shrub' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"])),'Gv']
        maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='shrub' & natopen.ref.cov.val$grunn==as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet_1til5000"]),'maxmin']
        # coercing x into results.natopen.GRUK dataframe
        results.natopen.GRUK[['scaled']][i,'shrub'] <- scal() 
        results.natopen.GRUK[['non-truncated']][i,'shrub'] <- scal() 
        results.natopen.GRUK[['original']][i,'shrub'] <- val
        
      }
    
  }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}

# for using both sides of the plant indicators
results.natopen.GRUK[['2-sided']] <- results.natopen.GRUK[['non-truncated']]

# remove values >1 for 2-sided indicators
results.natopen.GRUK[['2-sided']]$CC1[results.natopen.GRUK[['2-sided']]$CC1>1] <- NA
results.natopen.GRUK[['2-sided']]$CC2[results.natopen.GRUK[['2-sided']]$CC2>1] <- NA

results.natopen.GRUK[['2-sided']]$SS1[results.natopen.GRUK[['2-sided']]$SS1>1] <- NA
results.natopen.GRUK[['2-sided']]$SS2[results.natopen.GRUK[['2-sided']]$SS2>1] <- NA

results.natopen.GRUK[['2-sided']]$RR1[results.natopen.GRUK[['2-sided']]$RR1>1] <- NA
results.natopen.GRUK[['2-sided']]$RR2[results.natopen.GRUK[['2-sided']]$RR2>1] <- NA

results.natopen.GRUK[['2-sided']]$Light1[results.natopen.GRUK[['2-sided']]$Light1>1] <- NA
results.natopen.GRUK[['2-sided']]$Light2[results.natopen.GRUK[['2-sided']]$Light2>1] <- NA

results.natopen.GRUK[['2-sided']]$Nitrogen1[results.natopen.GRUK[['2-sided']]$Nitrogen1>1] <- NA
results.natopen.GRUK[['2-sided']]$Nitrogen2[results.natopen.GRUK[['2-sided']]$Nitrogen2>1] <- NA

results.natopen.GRUK[['2-sided']]$Soil_disturbance1[results.natopen.GRUK[['2-sided']]$Soil_disturbance1>1] <- NA
results.natopen.GRUK[['2-sided']]$Soil_disturbance2[results.natopen.GRUK[['2-sided']]$Soil_disturbance2>1] <- NA

#saveRDS(results.natopen.GRUK, "data/cache/results.natopen.GRUK.RDS")
```



```{r, include = F}

# Use cache data
results.natopen.GRUK<-readRDS(paste0(here::here(),"/data/cache/results.natopen.GRUK.RDS"))
```

```{r}
head(results.natopen.GRUK[['2-sided']])
```
#### Scaled value analyses

In order to visualize the results we need to rearrange the results-objects from wide to long format (note that there is both a lower and an upper condition indicator for each of the functional plant indicators).

```{r, message = FALSE}
#### scaled values by main ecosystem type ####
## continuing with 2-sided
res.natopen.GRUK <- results.natopen.GRUK[['2-sided']]

# make long version of the scaled value part
res.natopen.GRUK <-
  res.natopen.GRUK %>% 
  pivot_longer(
    cols = c("CC1","CC2",
             "SS1","SS2",
             "RR1","RR2",
             "Light1","Light2",
             "Nitrogen1","Nitrogen2",
             "Soil_disturbance1","Soil_disturbance2",
             "aliens","erosion","shrub"),
    names_to = "fp_ind",
    values_to = "scaled_value",
    values_drop_na = FALSE
  )

# add original values as well
res.natopen.GRUK <- 
  res.natopen.GRUK %>% add_column(original = results.natopen.GRUK[['original']] %>% 
                                   pivot_longer(
                                     cols = c("CC1","CC2",
                                              "SS1","SS2",
                                              "RR1","RR2",
                                              "Light1","Light2",
                                              "Nitrogen1","Nitrogen2",
                                              "Soil_disturbance1","Soil_disturbance2",
                                              "aliens","erosion","shrub"),
                                     names_to = NULL,
                                     values_to = "original",
                                     values_drop_na = FALSE
                                   ) %>%
                                   pull(original)
  )

summary(res.natopen.GRUK)
res.natopen.GRUK <- res.natopen.GRUK %>%
  mutate(Dekning_problemarter = as.numeric(Dekning_problemarter))
summary(res.natopen.GRUK)

```

### Ecologcial condition indicators
We can show the resulting scaled values as Violin plots for each the circle-based indicators
```{r,echo=FALSE,message=FALSE, warning=FALSE}

# plotting scaled values by indicator
res.natopen.GRUK %>%
  filter(fp_ind %in% c("aliens","erosion","shrub")) %>%
  ggplot(aes(x=fp_ind,y=scaled_value, fill=fp_ind)) +
  geom_hline(yintercept=0.6, linetype="dashed") + 
  geom_violin(color=NA)+
  geom_point(size=1, shape=16, color="black") + 
  xlab("Area without cover of...") + 
  ylab("Scaled indicator value") + 
  theme(legend.position="none")


```

We can show the resulting scaled values as Violin plots for each plant indicator and main ecosystem type 
```{r,echo=FALSE,message=FALSE, warning=FALSE}

# plotting scaled values by ecosystem type
ggplot(res.natopen.GRUK, aes(x=factor(Kartleggingsenhet_1til5000), y=scaled_value, fill=fp_ind)) + 
  geom_hline(yintercept=0.6, linetype="dashed") + 
  geom_violin(color = NA) +
  #  geom_boxplot(width=0.2, color="grey") +
  geom_point(size=1, shape=16, color="black") +
  facet_wrap(~factor(fp_ind,levels=c("CC1","SS1","RR1","Light1","Nitrogen1","Soil_disturbance1",
                                     "CC2","SS2","RR2","Light2","Nitrogen2","Soil_disturbance2"),
                     labels=c("CSR-C1","CSR-S1","CSR-R1","Light1","Nitrogen1","Soil_disturbance1",
                              "CSR-C2","CSR-S2","CSR-R2","Light2","Nitrogen2","Soil_disturbance2")), ncol = 6) + 
  xlab("Basic ecosystem type") + 
  ylab("Scaled indicator value (GRUK data, all indicators)") + 
  theme(legend.position="none") +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust=0.2))


```
The ANO results show a high frequency of too low CSR-R values (CSR-R1) and Soil disturbance while also showing signs of higher than expected competition (CSR-C2), which  indicates that a number of naturally open areas are changing in their dominance structure and physical stability. The GRUK data show a generally large spread of scaled values, but the violin plots (the shapes indicate the relative amount of observations across the y-axis) suggest that there is a number of sites with too high nitrogen (Nitrogen2), and some sites with too much disturbance (CSR-R2) and light (Light2). Many GRUK sites are popular outdoor locations with the local population, which could increase disturbance and thus cause deviations towards high CSR-R values (CSR-R2).
Increased nitrogen may be an effect of alien species, which the GRUK data records as a major pressure and problem for many monitoring sites.

```{r}

x <- c("erosjon_prosent","Dekning_busker_busksjikt","Dekning_tresjikt","Dekning_problemarter" ,"Totaldekning_fremmedearter")
y1 <- c("CC1","CC2","SS1","SS2","RR1","RR2")
y2 <- c("Light1","Light2","Nitrogen1","Nitrogen2","Soil_disturbance1","Soil_disturbance2")


par(mfrow=c(5,6))
for (i in x) {
  for (j in y1) {
    df <- res.natopen.GRUK[res.natopen.GRUK$fp_ind==j,c(i,'scaled_value')]
    names(df) <- c("xx","yy")
    df$xx <- as.numeric(df$xx)
    df$yy <- as.numeric(df$yy)
    df[!is.na(df$yy) & df$yy==1,"yy"] <- 0.9999
    df[!is.na(df$yy) & df$yy==0,"yy"] <- 0.0001
    with(df, plot(xx,yy,xlab=i,ylab=j, xlim=c(0,115), ylim=c(0,1)) )
    mod <- with(df, betareg(yy~xx))
    points(0:100,predict(mod,newdata=data.frame(xx=0:100),type='response'),col="red",type="l")
    text(110,0.1,labels=paste(round(summary(mod)$coefficients$mean[2,4],digits=3)), col="red")
  }
}

par(mfrow=c(5,6))
for (i in x) {
  for (j in y2) {
    df <- res.natopen.GRUK[res.natopen.GRUK$fp_ind==j,c(i,'scaled_value')]
    names(df) <- c("xx","yy")
    df$xx <- as.numeric(df$xx)
    df$yy <- as.numeric(df$yy)
    df[!is.na(df$yy) & df$yy==1,"yy"] <- 0.9999
    df[!is.na(df$yy) & df$yy==0,"yy"] <- 0.0001
    with(df, plot(xx,yy,xlab=i,ylab=j, xlim=c(0,115), ylim=c(0,1)) )
    mod <- with(df, betareg(yy~xx))
    points(0:100,predict(mod,newdata=data.frame(xx=0:100),type='response'),col="red",type="l")
    text(110,0.1,labels=paste(round(summary(mod)$coefficients$mean[2,4],digits=3)), col="red")
  }
}


```
From this overview we focus on negative trends (i.e. condition variables increase towards worse and condition indicators decrease towards worse).
  - too low light (Light1) may be a function of high tree cover.
  - too high light (Light2) may be a function of high erosion.
  - too high nitrogen (nitrogen2) may be a function of high alien cover.
  - too low disturbance (Soil_disturbance1) may be a function of high tree cover.
  - too high disturbance (Soil_disturbance2) may be a function of high problem species cover.
  - too high competition strategy (CC2), may be a function of both high cover of alien and problem species

```{r}
res.natopen.GRUK <- res.natopen.GRUK %>%
  mutate(scaled_value2 = ifelse(scaled_value < 0.001, 0.001, scaled_value) ) %>%
  mutate(scaled_value2 = ifelse(scaled_value2 > 0.999, 0.999, scaled_value2) )


summary( glmmTMB(scaled_value2 ~ Dekning_tresjikt +(1|LokalitetID), family=beta_family(), data=res.natopen.GRUK[res.natopen.GRUK$fp_ind=="Light1",]) )

summary( glmmTMB(scaled_value2 ~ erosjon_prosent +(1|LokalitetID), family=beta_family(), data=res.natopen.GRUK[res.natopen.GRUK$fp_ind=="Light2",]) )

summary( glmmTMB(scaled_value2 ~ Totaldekning_fremmedearter +(1|LokalitetID), family=beta_family(), data=res.natopen.GRUK[res.natopen.GRUK$fp_ind=="Nitrogen2",]) )

summary( glmmTMB(scaled_value2 ~ Dekning_tresjikt +(1|LokalitetID), family=beta_family(), data=res.natopen.GRUK[res.natopen.GRUK$fp_ind=="Soil_disturbance1",]) )

summary( glmmTMB(scaled_value2 ~ Dekning_problemarter +(1|LokalitetID), family=beta_family(), data=res.natopen.GRUK[res.natopen.GRUK$fp_ind=="Soil_disturbance2",]) )

summary( glmmTMB(scaled_value2 ~ Totaldekning_fremmedearter +(1|LokalitetID), family=beta_family(), data=res.natopen.GRUK[res.natopen.GRUK$fp_ind=="CC2",]) )

summary( glmmTMB(scaled_value2 ~ Dekning_problemarter +(1|LokalitetID), family=beta_family(), data=res.natopen.GRUK[res.natopen.GRUK$fp_ind=="CC2",]) )


```

Plotting relevant scatterplots and models
```{r}

## Light1 vs. Dekning_tresjikt
# basic scatterplot
p_L1_tre <- res.natopen.GRUK %>%
  filter(fp_ind=="Light1") %>%
  ggplot(aes(x=Dekning_tresjikt, y=scaled_value)) +
  geom_point() +
  coord_cartesian(xlim = c(0, 100),ylim = c(0, 1)) +
  labs(y = "Light1", x = "Tree cover (%)") # (Light index towards lower limit value)

# model and model prediction
mod <- glmmTMB(scaled_value2 ~ Dekning_tresjikt +(1|LokalitetID), family=beta_family(), data=res.natopen.GRUK[res.natopen.GRUK$fp_ind=="Light1",])
pred <- data.frame(
  mu = predict(mod,newdata=data.frame(Dekning_tresjikt=0:100),type='link',re.form=NA),
  se = predict(mod,newdata=data.frame(Dekning_tresjikt=0:100),type='link',re.form=NA,se.fit=T)$se.fit
)
pred <- pred %>%
  mutate(ci.lo = mu-1.96*se,
         ci.hi = mu+1.96*se)

# expit function for backtransforming from logit-link
expit <- function(L) exp(L) / (1+exp(L))

# add model line and 95%-CI to scatterplot
p_L1_tre <- p_L1_tre +
  geom_line(data=data.frame(x=0:100,y=expit(pred$mu)), aes(x,y),col="firebrick",linewidth=2) +
  geom_ribbon(data = pred, aes(x= 0:100, y = mu, ymin = expit(ci.lo), ymax = expit(ci.hi)), alpha = 0.3, fill = "firebrick") + 
#  annotate("text",x=90, y=0.1, label=paste("p = ",round(summary(mod)$coefficients$cond[2,4],6)), size=6, col="firebrick" ) +
  annotate("text",x=90, y=0.1, label=paste("p < 0.01"), size=6, col="firebrick" )


## Light2 vs. erosjon_prosent
# basic scatterplot
p_L2_slit <- res.natopen.GRUK %>%
  filter(fp_ind=="Light2") %>%
  ggplot(aes(x=erosjon_prosent, y=scaled_value)) +
  geom_point() +
  coord_cartesian(xlim = c(0, 100),ylim = c(0, 1)) +
  labs(y = "Light2", x = "Erosion/Disturbance cover (%)") # (Light index towards upper limit value)

# model and model prediction
mod <- glmmTMB(scaled_value2 ~ erosjon_prosent +(1|LokalitetID), family=beta_family(), data=res.natopen.GRUK[res.natopen.GRUK$fp_ind=="Light2",])
pred <- data.frame(
  mu = predict(mod,newdata=data.frame(erosjon_prosent=0:100),type='link',re.form=NA),
  se = predict(mod,newdata=data.frame(erosjon_prosent=0:100),type='link',re.form=NA,se.fit=T)$se.fit
)
pred <- pred %>%
  mutate(ci.lo = mu-1.96*se,
         ci.hi = mu+1.96*se)

# expit function for backtransforming from logit-link
expit <- function(L) exp(L) / (1+exp(L))

# add model line and 95%-CI to scatterplot
p_L2_slit <- p_L2_slit +
  geom_line(data=data.frame(x=0:100,y=expit(pred$mu)), aes(x,y),col="firebrick",linewidth=2) +
  geom_ribbon(data = pred, aes(x= 0:100, y = mu, ymin = expit(ci.lo), ymax = expit(ci.hi)), alpha = 0.3, fill = "firebrick") + 
#  annotate("text",x=90, y=0.1, label=paste("p = ",round(summary(mod)$coefficients$cond[2,4],6)), size=6, col="firebrick" ) +
  annotate("text",x=90, y=0.1, label=paste("p < 0.01"), size=6, col="firebrick" )


## Nitrogen2 vs. Totaldekning_fremmedearter
# basic scatterplot
p_N2_frem <- res.natopen.GRUK %>%
  filter(fp_ind=="Nitrogen2") %>%
  ggplot(aes(x=Totaldekning_fremmedearter, y=scaled_value)) +
  geom_point() +
  coord_cartesian(xlim = c(0, 100),ylim = c(0, 1)) +
  labs(y = "Nitrogen2", x = "Alien plant species cover (%)") # (Nitrogen index towards upper limit value)

# model and model prediction
mod <- glmmTMB(scaled_value2 ~ Totaldekning_fremmedearter +(1|LokalitetID), family=beta_family(), data=res.natopen.GRUK[res.natopen.GRUK$fp_ind=="Nitrogen2",])
pred <- data.frame(
  mu = predict(mod,newdata=data.frame(Totaldekning_fremmedearter=0:100),type='link',re.form=NA),
  se = predict(mod,newdata=data.frame(Totaldekning_fremmedearter=0:100),type='link',re.form=NA,se.fit=T)$se.fit
)
pred <- pred %>%
  mutate(ci.lo = mu-1.96*se,
         ci.hi = mu+1.96*se)

# expit function for backtransforming from logit-link
expit <- function(L) exp(L) / (1+exp(L))

# add model line and 95%-CI to scatterplot
p_N2_frem <- p_N2_frem +
  geom_line(data=data.frame(x=0:100,y=expit(pred$mu)), aes(x,y),col="firebrick",linewidth=2) +
  geom_ribbon(data = pred, aes(x= 0:100, y = mu, ymin = expit(ci.lo), ymax = expit(ci.hi)), alpha = 0.3, fill = "firebrick") + 
#  annotate("text",x=90, y=0.1, label=paste("p = ",round(summary(mod)$coefficients$cond[2,4],6)), size=6, col="firebrick" ) +
  annotate("text",x=90, y=0.1, label=paste("p < 0.01"), size=6, col="firebrick" )


## Soil_disturbance1 vs. Dekning_tresjikt
# basic scatterplot
p_Sd1_tre <- res.natopen.GRUK %>%
  filter(fp_ind=="Soil_disturbance1") %>%
  ggplot(aes(x=Dekning_tresjikt, y=scaled_value)) +
  geom_point() +
  coord_cartesian(xlim = c(0, 100),ylim = c(0, 1)) +
  labs(y = "Disturbance1", x = "Tree cover (%)") # (Soil disturbance index towards lower limit value)

# model and model prediction
mod <- glmmTMB(scaled_value2 ~ Dekning_tresjikt +(1|LokalitetID), family=beta_family(), data=res.natopen.GRUK[res.natopen.GRUK$fp_ind=="Soil_disturbance1",])
pred <- data.frame(
  mu = predict(mod,newdata=data.frame(Dekning_tresjikt=0:100),type='link',re.form=NA),
  se = predict(mod,newdata=data.frame(Dekning_tresjikt=0:100),type='link',re.form=NA,se.fit=T)$se.fit
)
pred <- pred %>%
  mutate(ci.lo = mu-1.96*se,
         ci.hi = mu+1.96*se)

# expit function for backtransforming from logit-link
expit <- function(L) exp(L) / (1+exp(L))

# add model line and 95%-CI to scatterplot
p_Sd1_tre <- p_Sd1_tre +
  geom_line(data=data.frame(x=0:100,y=expit(pred$mu)), aes(x,y),col="firebrick",linewidth=2) +
  geom_ribbon(data = pred, aes(x= 0:100, y = mu, ymin = expit(ci.lo), ymax = expit(ci.hi)), alpha = 0.3, fill = "firebrick") + 
#  annotate("text",x=90, y=0.1, label=paste("p = ",round(summary(mod)$coefficients$cond[2,4],6)), size=6, col="firebrick" ) +
  annotate("text",x=90, y=0.1, label=paste("p < 0.01"), size=6, col="firebrick" )


## Soil_disturbance2 vs. Dekning_problemarter
# basic scatterplot
p_Sd2_prob <- res.natopen.GRUK %>%
  filter(fp_ind=="Soil_disturbance2") %>%
  ggplot(aes(x=Dekning_problemarter, y=scaled_value)) +
  geom_point() +
  coord_cartesian(xlim = c(0, 100),ylim = c(0, 1)) +
  labs(y = "Disturbance2", x = "Problem species cover (%)") # (Soil disturbance index towards upper limit value)

# model and model prediction
mod <- glmmTMB(scaled_value2 ~ Dekning_problemarter +(1|LokalitetID), family=beta_family(), data=res.natopen.GRUK[res.natopen.GRUK$fp_ind=="Soil_disturbance2",])
pred <- data.frame(
  mu = predict(mod,newdata=data.frame(Dekning_problemarter=0:100),type='link',re.form=NA),
  se = predict(mod,newdata=data.frame(Dekning_problemarter=0:100),type='link',re.form=NA,se.fit=T)$se.fit
)
pred <- pred %>%
  mutate(ci.lo = mu-1.96*se,
         ci.hi = mu+1.96*se)

# expit function for backtransforming from logit-link
expit <- function(L) exp(L) / (1+exp(L))

# add model line and 95%-CI to scatterplot
p_Sd2_prob <- p_Sd2_prob +
  geom_line(data=data.frame(x=0:100,y=expit(pred$mu)), aes(x,y),col="firebrick",linewidth=2) +
  geom_ribbon(data = pred, aes(x= 0:100, y = mu, ymin = expit(ci.lo), ymax = expit(ci.hi)), alpha = 0.3, fill = "firebrick") + 
#  annotate("text",x=90, y=0.1, label=paste("p = ",round(summary(mod)$coefficients$cond[2,4],6)), size=6, col="firebrick" ) +
  annotate("text",x=90, y=0.1, label=paste("p < 0.01"), size=6, col="firebrick" )


## CC2 vs. Dekning_problemarter
# basic scatterplot
p_CC2_prob <- res.natopen.GRUK %>%
  filter(fp_ind=="CC2") %>%
  ggplot(aes(x=Dekning_problemarter, y=scaled_value)) +
  geom_point() +
  coord_cartesian(xlim = c(0, 100),ylim = c(0, 1)) +
  labs(y = "CSR-C2", x = "Problem species cover (%)") #  (Grime's C index towards upper limit value)

# model and model prediction
mod <- glmmTMB(scaled_value2 ~ Dekning_problemarter +(1|LokalitetID), family=beta_family(), data=res.natopen.GRUK[res.natopen.GRUK$fp_ind=="CC2",])
pred <- data.frame(
  mu = predict(mod,newdata=data.frame(Dekning_problemarter=0:100),type='link',re.form=NA),
  se = predict(mod,newdata=data.frame(Dekning_problemarter=0:100),type='link',re.form=NA,se.fit=T)$se.fit
)
pred <- pred %>%
  mutate(ci.lo = mu-1.96*se,
         ci.hi = mu+1.96*se)

# expit function for backtransforming from logit-link
expit <- function(L) exp(L) / (1+exp(L))

# add model line and 95%-CI to scatterplot
p_CC2_prob <- p_CC2_prob +
  geom_line(data=data.frame(x=0:100,y=expit(pred$mu)), aes(x,y),col="firebrick",linewidth=2) +
  geom_ribbon(data = pred, aes(x= 0:100, y = mu, ymin = expit(ci.lo), ymax = expit(ci.hi)), alpha = 0.3, fill = "firebrick") + 
  annotate("text",x=90, y=0.1, label=paste("p = ",round(summary(mod)$coefficients$cond[2,4],2)), size=6, col="firebrick" ) #+
#  annotate("text",x=90, y=0.1, label=paste("p < 0.01"), size=6, col="firebrick" )


## CC2 vs. Totaldekning_fremmedearter
# basic scatterplot
p_CC2_frem <- res.natopen.GRUK %>%
  filter(fp_ind=="CC2") %>%
  ggplot(aes(x=Totaldekning_fremmedearter, y=scaled_value)) +
  geom_point() +
  coord_cartesian(xlim = c(0, 100),ylim = c(0, 1)) +
  labs(y = "CSR-C2", x = "Alien species cover (%)") # (Grime's C index towards upper limit value)

# model and model prediction
mod <- glmmTMB(scaled_value2 ~ Totaldekning_fremmedearter +(1|LokalitetID), family=beta_family(), data=res.natopen.GRUK[res.natopen.GRUK$fp_ind=="CC2",])
pred <- data.frame(
  mu = predict(mod,newdata=data.frame(Totaldekning_fremmedearter=0:100),type='link',re.form=NA),
  se = predict(mod,newdata=data.frame(Totaldekning_fremmedearter=0:100),type='link',re.form=NA,se.fit=T)$se.fit
)
pred <- pred %>%
  mutate(ci.lo = mu-1.96*se,
         ci.hi = mu+1.96*se)

# expit function for backtransforming from logit-link
expit <- function(L) exp(L) / (1+exp(L))

# add model line and 95%-CI to scatterplot
p_CC2_frem <- p_CC2_frem +
  geom_line(data=data.frame(x=0:100,y=expit(pred$mu)), aes(x,y),col="firebrick",linewidth=2) +
  geom_ribbon(data = pred, aes(x= 0:100, y = mu, ymin = expit(ci.lo), ymax = expit(ci.hi)), alpha = 0.3, fill = "firebrick") + 
  annotate("text",x=90, y=0.1, label=paste("p = ",round(summary(mod)$coefficients$cond[2,4],2)), size=6, col="firebrick" ) #+
#  annotate("text",x=90, y=0.1, label=paste("p < 0.01"), size=6, col="firebrick" )


ggarrange(p_L1_tre, p_L2_slit, p_Sd1_tre, p_Sd2_prob, p_CC2_prob, p_CC2_frem, p_N2_frem + rremove("x.text"), 
          labels = c("A", "B", "C", "D", "E", "F", "G"),
          ncol = 2, nrow = 4)



```


### maps
We can also show the results as a map, for instance for CSR-R1 (the lower ruderal/disturbance indicator) for ANO and Nitrogen2 (the upper nitrogen indicator) for GRUK, either by directly plotting the data onto the map...
```{r, echo=FALSE, warning=FALSE,message = FALSE}
#### scaled value maps ####
# keep wide format and add geometry again
res.natopen.GRUK2 <- results.natopen.GRUK[['2-sided']]
res.natopen.GRUK2 <- st_as_sf(res.natopen.GRUK2, coords = c("UTM33_E","UTM33_N"), remove=F, crs = 25833)

## Import Norway map
nor <- st_read(paste0(here::here(),"/data/mapdata/Norway outline/outlineOfNorway_EPSG25833.shp")) #%>%
#  st_as_sf() %>%
#  st_transform(crs = st_crs(myr))

## Import region map
reg <- st_read(paste0(here::here(),"/data/mapdata/Norway outline/regions.shp")) #%>%
#  st_as_sf() %>%
#  st_transform(crs = st_crs(ANO.geo))

# change region names to something R-friendly
#reg$region
reg$region <- c("Northern Norway","Central Norway","Eastern Norway","Western Norway","Southern Norway")

regnor <- st_intersection(reg,nor)

# Bounding boxes for GRUK area
boks <- st_bbox(c(xmin = 194100, xmax = 263300, ymax = 6649700, ymin = 6549000), crs = st_crs(25833))

boks.Oslo <- st_bbox(c(xmin = 247100, xmax = 263300, ymax = 6649700, ymin = 6628000), crs = st_crs(25833))
boks.mid <- st_bbox(c(xmin = 234000, xmax = 257300, ymax = 6621000, ymin = 6590000), crs = st_crs(25833))
boks.south <- st_bbox(c(xmin = 194100, xmax = 204800, ymax = 6568700, ymin = 6545000), crs = st_crs(25833))

# fixing some variable types - results
res.natopen.GRUK2$Lokalitetskvalitet <- as.factor(res.natopen.GRUK2$Lokalitetskvalitet)
res.natopen.GRUK2$Tilstand <- as.factor(res.natopen.GRUK2$Tilstand)
res.natopen.GRUK2$Naturmangfold <- as.factor(res.natopen.GRUK2$Naturmangfold)

res.natopen.GRUK2$Lokalitetskvalitet <- factor(res.natopen.GRUK2$Lokalitetskvalitet,levels=c("Lav kvalitet","Moderat kvalitet","Høy kvalitet","Svært høy kvalitet"))
res.natopen.GRUK2$Tilstand <- factor(res.natopen.GRUK2$Tilstand, levels=c("Dårlig","Moderat","God"))
res.natopen.GRUK2$Naturmangfold <- factor(res.natopen.GRUK2$Naturmangfold,levels=c("Lite","Moderat","Stort"))

# fixing some variable types - polygon data
#summary(GRUK.polygoner)
GRUK.polygoner2 <- GRUK.polygoner %>%
  mutate(Tilstand = factor(Tilstand, levels=c("Svært redusert","Dårlig","Moderat","God"))) %>%
  mutate(Naturmangfold = factor(Naturmangfold, levels=c("0","Lite","Moderat","Stort"))) %>%
  mutate(Lokalitetskvalitet = factor(Lokalitetskvalitet, levels=c("Lav kvalitet","Moderat kvalitet","Høy kvalitet","Svært høy kvalitet","Ikke kvalitetsvurdert"))) %>%
  filter(!is.na(Tilstand))
summary(GRUK.polygoner2[,c("Lokalitetskvalitet","Naturmangfold","Tilstand")])[,1:3]

# creating spatial objects for the bounding boxes
bokses <- data.frame(lon=c(241100,241100,270300,270300,
                           225000,225000,270300,270300,
                           184100,184100,214800,214800),
                     lat=c(6628000,6649700,6649700,6628000,
                           6590000,6621000,6621000,6590000,
                           6545000,6568700,6568700,6545000),
                     omr=c(rep('indre',4),rep('midtre',4),rep('ytre',4))
)

bokses <- bokses %>%
  st_as_sf(coords = c("lon", "lat"), remove=F, crs = 25833) %>%
  group_by(omr) %>%
  summarise(geometry = st_combine(geometry)) %>%
  st_cast("POLYGON")

```


## condition evaluation, maps
```{r}
map.all <- tm_shape(regnor, bbox=boks) +
  tm_fill('GID_0', labels="", title="", legend.show = FALSE) + 
  tm_borders() +
  tm_graticules() +
  tm_shape(bokses) +
  tm_borders(lwd=3) +
  tm_shape(GRUK.polygoner2) +
  tm_dots('Tilstand',palette=tmaptools::get_brewer_pal("RdYlGn", 4, plot = FALSE), scale=2, legend.show = FALSE) +
  tm_add_legend(type = "fill", 
                col = tmaptools::get_brewer_pal("RdYlGn", 4, plot = FALSE),
                labels = c("Svært redusert", "Dårlig", "Moderat", "God"),
                title = "") +  #
  tm_layout(main.title = "Tilstand GRUK lokaliteter (Mdir instruks)", main.title.size=1.2, legend.position = c("left", "top"), legend.text.size = 1, legend.title.size=1.5)

map.indre <- tm_shape(regnor, bbox=boks.Oslo) +
  tm_fill('GID_0', labels="", title="", legend.show = FALSE) + 
  tm_borders() +
  tm_graticules() +
  tm_shape(GRUK.polygoner2) +
  tm_dots('Tilstand',palette=tmaptools::get_brewer_pal("RdYlGn", 3, plot = FALSE), scale=4, legend.show = FALSE) +  
  tm_layout(main.title = "Indre Oslofjord",legend.position = c("right", "bottom"), main.title.size=1.2)

map.midtre <- tm_shape(regnor, bbox=boks.mid) +
  tm_fill('GID_0', labels="", title="", legend.show = FALSE) + 
  tm_borders() +
  tm_graticules() +
  tm_shape(GRUK.polygoner2) +
  tm_dots('Tilstand',palette=tmaptools::get_brewer_pal("RdYlGn", 3, plot = FALSE), scale=4, legend.show = FALSE) +  
  tm_layout(main.title = "Midtre Oslofjord",legend.position = c("right", "bottom"), main.title.size=1.2)

map.ytre <- tm_shape(regnor, bbox=boks.south) +
  tm_fill('GID_0', labels="", title="", legend.show = FALSE) + 
  tm_borders() +
  tm_graticules() +
  tm_shape(GRUK.polygoner2) +
  tm_dots('Tilstand',palette=tmaptools::get_brewer_pal("RdYlGn", 3, plot = FALSE), scale=4, legend.show = FALSE) +  
  tm_layout(main.title = "Ytre Oslofjord",legend.position = c("right", "bottom"), main.title.size=1.2)

tmap_arrange(map.all, map.indre, map.ytre, map.midtre, nrow=2)

```
## biodiversity evaluation, maps
```{r}
tm_shape(regnor, bbox=boks) +
  tm_fill('GID_0', labels="", title="", legend.show = FALSE) + 
  tm_borders() +
  tm_graticules() +
  tm_shape(res.natopen.GRUK2) +
  tm_dots('Naturmangfold',palette=tmaptools::get_brewer_pal("RdYlGn", 3, plot = FALSE), scale=4, legend.show = TRUE) +  
  tm_layout(main.title = "Naturmangfold, GRUK lokaliteter",legend.position = c("right", "bottom"), main.title.size=1.2)


tm_shape(regnor, bbox=boks.Oslo) +
  tm_fill('GID_0', labels="", title="", legend.show = FALSE) + 
  tm_borders() +
  tm_graticules() +
  tm_shape(res.natopen.GRUK2) +
  tm_dots('Naturmangfold',palette=tmaptools::get_brewer_pal("RdYlGn", 3, plot = FALSE), scale=4, legend.show = TRUE) +  
  tm_layout(main.title = "Naturmangfold, GRUK lokaliteter Oslo",legend.position = c("right", "bottom"), main.title.size=1.2)

tm_shape(regnor, bbox=boks.mid) +
  tm_fill('GID_0', labels="", title="", legend.show = FALSE) + 
  tm_borders() +
  tm_graticules() +
  tm_shape(res.natopen.GRUK2) +
  tm_dots('Naturmangfold',palette=tmaptools::get_brewer_pal("RdYlGn", 3, plot = FALSE), scale=4, legend.show = TRUE) +  
  tm_layout(main.title = "Naturmangfold, GRUK lokaliteter Oslofjord",legend.position = c("right", "bottom"), main.title.size=1.2)

tm_shape(regnor, bbox=boks.south) +
  tm_fill('GID_0', labels="", title="", legend.show = FALSE) + 
  tm_borders() +
  tm_graticules() +
  tm_shape(res.natopen.GRUK2) +
  tm_dots('Naturmangfold',palette=tmaptools::get_brewer_pal("RdYlGn", 3, plot = FALSE), scale=4, legend.show = TRUE) +  
  tm_layout(main.title = "Naturmangfold, GRUK lokaliteter Langesundsfjorden",legend.position = c("right", "bottom"), main.title.size=1.2)


```
## site quality evaluation, maps
```{r}
tm_shape(regnor, bbox=boks) +
  tm_fill('GID_0', labels="", title="", legend.show = FALSE) + 
  tm_borders() +
  tm_graticules() +
  tm_shape(res.natopen.GRUK2) +
  tm_dots('Lokalitetskvalitet',palette=tmaptools::get_brewer_pal("RdYlGn", 3, plot = FALSE), scale=4, legend.show = TRUE) +  
  tm_layout(main.title = "Lokalitetskvalitet, GRUK lokaliteter",legend.position = c("right", "bottom"), main.title.size=1.2)


tm_shape(regnor, bbox=boks.Oslo) +
  tm_fill('GID_0', labels="", title="", legend.show = FALSE) + 
  tm_borders() +
  tm_graticules() +
  tm_shape(res.natopen.GRUK2) +
  tm_dots('Lokalitetskvalitet',palette=tmaptools::get_brewer_pal("RdYlGn", 3, plot = FALSE), scale=4, legend.show = TRUE) +  
  tm_layout(main.title = "Lokalitetskvalitet, GRUK lokaliteter Oslo",legend.position = c("right", "bottom"), main.title.size=1.2)

tm_shape(regnor, bbox=boks.mid) +
  tm_fill('GID_0', labels="", title="", legend.show = FALSE) + 
  tm_borders() +
  tm_graticules() +
  tm_shape(res.natopen.GRUK2) +
  tm_dots('Lokalitetskvalitet',palette=tmaptools::get_brewer_pal("RdYlGn", 3, plot = FALSE), scale=4, legend.show = TRUE) +  
  tm_layout(main.title = "Lokalitetskvalitet, GRUK lokaliteter Oslofjord",legend.position = c("right", "bottom"), main.title.size=1.2)

tm_shape(regnor, bbox=boks.south) +
  tm_fill('GID_0', labels="", title="", legend.show = FALSE) + 
  tm_borders() +
  tm_graticules() +
  tm_shape(res.natopen.GRUK2) +
  tm_dots('Lokalitetskvalitet',palette=tmaptools::get_brewer_pal("RdYlGn", 3, plot = FALSE), scale=4, legend.show = TRUE) +  
  tm_layout(main.title = "Lokalitetskvalitet, GRUK lokaliteter Langesundsfjorden",legend.position = c("right", "bottom"), main.title.size=1.2)


```

## scaled value maps
```{r}

# GRUK, Light2 (upper indicator)
tm_shape(regnor) +
  tm_fill('GID_0', labels="", title="", legend.show = FALSE) + 
  tm_borders() +
  tm_graticules() +
  tm_shape(res.natopen.GRUK2) +
  tm_dots('Light2',midpoint=NA, palette=tmaptools::get_brewer_pal("RdYlGn", 7, plot = FALSE), scale=2, legend.show = FALSE) + # 
  tm_layout(main.title = "Light index (upper), natopen GRUK",legend.position = c("right", "bottom"), main.title.size=1.2) + 
  tm_add_legend(type = "fill", 
                col = c(tmaptools::get_brewer_pal("RdYlGn", 7, plot = FALSE),'grey'),
                labels = c("0.3 - 0.4", "0.4 - 0.5", "0.5 - 0.6", "0.6 - 0.7", 
                           "0.7 - 0.8", "0.8 - 0.9", "0.9 - 1.0", ">1"),
                title = "index values")


```
For GRUK we can zoom in on the area around the Oslofjord...

```{r}
boks <- st_bbox(c(xmin = 194100, xmax = 263300, ymax = 6649700, ymin = 6549000), crs = st_crs(25833))

boks.Oslo <- st_bbox(c(xmin = 247100, xmax = 263300, ymax = 6649700, ymin = 6628000), crs = st_crs(25833))

tm_shape(regnor, bbox=boks) +
  tm_fill('GID_0', labels="", title="", legend.show = FALSE) + 
  tm_borders() +
  tm_graticules() +
  tm_shape(res.natopen.GRUK2) +
  tm_dots('Light2',midpoint=NA, palette=tmaptools::get_brewer_pal("RdYlGn", 7, plot = FALSE), scale=3, legend.show = FALSE) + # 
  tm_layout(main.title = "Light index (upper), natopen GRUK",legend.position = c("right", "bottom"), main.title.size=1.2)# + 
#  tm_add_legend(type = "fill", 
#                col = c(tmaptools::get_brewer_pal("RdYlGn", 7, plot = FALSE),'grey'),
#                labels = c("0.3 - 0.4", "0.4 - 0.5", "0.5 - 0.6", "0.6 - 0.7", 
#                           "0.7 - 0.8", "0.8 - 0.9", "0.9 - 1.0", ">1"),
#                title = "index values")

tm_shape(regnor, bbox=boks.Oslo) +
  tm_fill('GID_0', labels="", title="", legend.show = FALSE) + 
  tm_borders() +
  tm_graticules() +
  tm_shape(res.natopen.GRUK2) +
  tm_dots('Light2',midpoint=NA, palette=tmaptools::get_brewer_pal("RdYlGn", 7, plot = FALSE), scale=3, legend.show = FALSE) + # 
  tm_layout(main.title = "Light index (upper), natopen GRUK",legend.position = c("right", "bottom"), main.title.size=1.2) + 
  tm_add_legend(type = "fill", 
                col = c(tmaptools::get_brewer_pal("RdYlGn", 7, plot = FALSE),'grey'),
                labels = c("0.3 - 0.4", "0.4 - 0.5", "0.5 - 0.6", "0.6 - 0.7", 
                           "0.7 - 0.8", "0.8 - 0.9", "0.9 - 1.0", ">1"),
                title = "index values")

#YlOrRd
```
